---
title: "peridice_targeted_analysis"
author: "Iris Kern"
date: "2023-08-08"
output: html_document
---

# Setup
## Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(RColorBrewer)
# library(vegan)
```

## Data
raw peak list, ingalls standards, filtration and pcpn data
```{r}
# raw peak list
peak_list_neg <- read_csv('../csvs/neg_transition_reuslts.csv') %>% 
  mutate(`Molecule List Name` = str_replace(`Molecule List Name`,
                                            "HILIC_Neg_Standards", 
                                            "HILIC")) %>% 
  rename(column = `Molecule List Name`) %>% 
  rename(metabolite = Molecule) %>% 
  mutate(replicate_name = str_extract(`Replicate Name`, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_extract(`Replicate Name`, "27June|30June|14July|21July|27July")) %>%
  rename(replicate = `Replicate Name`) %>% 
  rename(area = Area) %>% 
  rename(background = Background) %>% 
  rename(precursor_mz = `Precursor Mz`) %>%
  rename(precursor_charge = `Precursor Charge`) %>%
  rename(product_mz = `Product Mz`) %>%
  rename(product_charge = `Product Charge`) %>%
  rename(frag_ion = `Fragment Ion`) %>%
  rename(rt = `Retention Time`) %>%
  rename(peak_rank = `Peak Rank`) %>%
  mutate(num_area = as.numeric(area)) %>% 
  mutate(num_background = as.numeric(background)) %>% 
  select(-c(precursor_charge, frag_ion, peak_rank)) %>%
  mutate(rt = as.numeric(as.factor(rt)))

peak_list_pos <- read_csv('../csvs/targeted_peak_list.csv') %>% 
  mutate(Protein = str_replace(Protein,
                                            "HILIC_Pos_Standards", 
                                            "HILIC")) %>% 
  rename(column = Protein) %>% 
  rename(metabolite = Peptide) %>% 
  mutate(replicate_name = str_extract(Replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_extract(Replicate, "27June|30June|14July|21July|27July")) %>% 
  rename(area = Area) %>% 
  rename(replicate = Replicate) %>% 
  rename(background = Background) %>% 
  rename(precursor_mz = `Precursor Mz`) %>%
  rename(precursor_charge = `Precursor Charge`) %>%
  rename(product_mz = `Product Mz`) %>%
  rename(product_charge = `Product Charge`) %>%
  rename(frag_ion = `Fragment Ion`) %>%
  rename(rt = `Retention Time`) %>%
  rename(peak_rank = `Peak Rank`) %>%
  mutate(num_area = as.numeric(area)) %>% 
  mutate(num_background = as.numeric(background)) %>% 
  select(-c(precursor_charge, frag_ion, peak_rank)) %>%
  mutate(rt = as.numeric(as.factor(rt)))

# sucrose <- peak_list_neg %>% 
#   filter(str_detect(metabolite, "Sucrose")) %>% 
#   filter(str_detect(replicate, "27July-RH_1|Poo|Mix|H2O"))
# 
# # bmised area
# sucrose_is <- peak_list_neg %>% 
#   na.omit() %>% 
#   filter(str_detect(metabolite, "Sucrose, 13C")) %>% 
#   mutate(mean_area = mean(num_area))

# rf ratio = 3628660

peak_list <- peak_list_neg %>% 
  rbind(peak_list_pos) %>% 
  drop_na(num_area)

## peak_list <- peak_list_pos

# standards
ingalls_standards <- read.csv(
  "https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv",
                              stringsAsFactors = FALSE, header = TRUE) %>%
  # filter(Column == Column.Type) %>%
  rename(internal_standards = Compound_Name) %>%
  select(internal_standards, Compound_Type, Concentration_uM, HILIC_Mix, Empirical_Formula) %>% #QE.RF.ratio
  filter(!is.na(Concentration_uM)) %>%
  filter(Compound_Type == "Internal Standard") %>%
  mutate(conc_nm = Concentration_uM * 1000) %>%
  select(-c(Concentration_uM)) %>%
  # rename(internal_standards = Metabolite.Name) %>%
  mutate(matched_cmpds = internal_standards) %>% 
  mutate(matched_cmpds = str_remove(matched_cmpds, ",.*")) %>% 
  select(-c(HILIC_Mix, Compound_Type, Empirical_Formula))

#filt data 
filt_data <- read_csv("../csvs/filt_data.csv") %>% 
  mutate(timepoint = str_replace_all(timepoint, "14July", "7-14-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "21July", "7-21-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "30June", "6-30-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "27July", "7-27-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "27June", "6-27-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%m-%d-%y")) %>% 
  select(-c("timepoint", "samp_type"))

# pcpn
pcpn <- read_xlsx("../metadata/peridice_pcpn.xlsx") %>% 
  select(Tank, Treatment2, Date, `PN (uM)`, `PC (uM)`, Cnratio, AccN, AccC, AddNn, AddN) %>% 
  rename(tank = Tank, treatment = Treatment2, date = Date, pn_um = `PN (uM)`, pc_um = `PC (uM)`, cn_ratio = Cnratio, 
         acc_n = AccN, acc_c = AccC, add_nn = AddNn, add_n = AddN) %>% 
  mutate(tank = str_remove(tank, "/(2|3)")) %>% 
  mutate(replicate = str_extract(tank, "\\d$")) %>% 
  mutate(replicate = as.numeric(replicate)) %>% 
  select(-c("tank")) %>% 
  group_by(treatment, date, replicate) %>% 
  mutate(pc_um = mean(pc_um)) %>% 
  mutate(pn_um = mean(pn_um)) %>% 
  mutate(cn_ratio = mean(cn_ratio)) %>% 
  mutate(acc_c = mean(acc_c)) %>% 
  mutate(acc_n = mean(acc_n)) %>% 
  distinct()

# add_n is nitrate add_nn is nitrate_nitrite

# setup metadata
metadata_targ <- filt_data %>% 
  left_join(pcpn, by = c("date", "replicate", "treatment")) %>% 
  na.omit() %>% 
  select(-c("replicate")) %>% 
  rename(replicate = filename) %>% 
  mutate(replicate = str_remove(replicate, ".mzML")) %>% 
  distinct()

metadataframe <- peak_list %>%
  # Grab just the unique filenames and rename the column
  distinct(replicate) %>% 
  rename(filename = replicate) %>% 
  mutate(filename = str_remove(filename, "230616_")) %>% 
  # Extract treatment information from the filename
  # Longest one needs to go first!")) %>%
  # Create a new column with timepoint information
  mutate(treatment = str_extract(filename, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote1|Tote2|Tote3|Tote4")) %>% 
  mutate(timepoint=str_extract(filename, "27June|30June|14July|21July|27July")) %>%
  mutate(replicate = str_extract(filename, "\\d$")) %>% 
  mutate(replicate = str_remove(replicate, ".mzML")) %>% 
  # Create a new column with sample type information
  mutate(samp_type=str_extract(filename, "Blk|Smp|Std|Poo")) %>%
  #mutate(treatment=ifelse(samp_type=="Poo", NA, treatment)) %>% 
  mutate(colid = levels(treatment)) %>% 
  distinct()

# parameters
area_min   <- 1 # peak area minimum - calculated from peak intensity of chromatogram
rt_flex    <- 0.4 # retention time flexibility - accounts variance across matrices and time on machine
blk_thresh <- 0.3 # blank sample threshold to calculate difference for samples
sn_min     <- 4 # signal to noise ratio minimum
```

# Quality Control
```{r}
peak_data_qc <- peak_list %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(replicate = str_remove(replicate, "230616_")) %>% 
  left_join(metadata_targ, by = c("replicate")) %>% 
  mutate(area_per_pc = num_area/pc_um)
```


# BMIS
## setup
```{r}
# Need to find minimum and maximum retention time for each sample
rt_table <- peak_list %>%
  filter(str_detect(replicate, regex("Std", ignore_case = TRUE))) %>%
  group_by(metabolite) %>%
  mutate(rt_min = min(rt, na.rm = TRUE)) %>%
  mutate(rt_max = max(rt, na.rm = TRUE)) %>%
  mutate(rt_reference = mean(rt, na.rm = TRUE)) %>%
  select(metabolite, rt_min, rt_max) %>%
  unique()

area_table <- peak_list %>%
  select(replicate, metabolite, num_area) %>%
  filter(str_detect(replicate, regex("Smp|Poo", ignore_case = TRUE)))

sn_table <- peak_list %>%
  filter(str_detect(replicate, regex("Smp|Poo", ignore_case = TRUE))) %>%
  select(replicate, metabolite, num_area, num_background) %>%
  mutate(signal_to_noise = (num_area / num_background))

blank_table <- peak_list %>%
    filter(str_detect(replicate_name, regex("Blk", ignore_case = TRUE))) %>%
    select(metabolite, num_area) %>%
    group_by(metabolite) %>% 
    filter(num_area == max(num_area)) %>%
    rename(blank_max = num_area) %>%
    unique()
```

## add flags for retention time, area, blanks, signal to noise
peak_list_all_flags
```{r}
peak_list_all_flags <- peak_list %>%
  left_join(rt_table) %>%
  mutate(rt_flag = ifelse((rt >= (rt_max + rt_flex) | rt <= (rt_min - rt_flex)), 
                          "RT.Flag", NA)) %>%
  left_join(blank_table) %>%
  group_by(metabolite) %>%
  mutate(blank_ref = num_area / blank_max) %>%
  mutate(is_present = ifelse(str_detect(replicate_name, ","), "Internal Std", "Non IS")) %>%
  mutate(blank_flag = ifelse(((is_present != "Internal Std") & ((num_area / blank_max) < blk_thresh)), 
                             "blank.Flag", 
                             ifelse(((is_present == "Internal Std") & ((num_area / blank_max) < blk_thresh)),
                                    "IS.blank.Flag", NA))) %>%
  mutate(area_min_flag = ifelse((num_area < area_min), "area_min_Flag", NA)) %>%
  mutate(area_with_qc   = ifelse(is.na(area_min_flag), num_area, NA)) %>%
  select(replicate:num_area, area_with_qc, everything()) %>%
  left_join(sn_table) %>%
  mutate(sn_flag = ifelse((signal_to_noise < sn_min), "sn_flag", NA)) %>% 
  filter_at(vars(blank_flag, area_min_flag, sn_flag),all_vars(is.na(.))) %>% 
  filter(!str_detect(area, "#N/A"))
```

## load in internal standards, generate qc_data, sampkey, and filtered out is areas vs non is areas
```{r}
internal_standards <- read.csv("https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv",
                               stringsAsFactors = FALSE, header = TRUE) %>%
  filter(Compound_Type == "Internal Standard")

qc_data <- peak_list_all_flags %>% 
  filter(!str_detect(replicate, "Blk|Std"))

sampkey <- qc_data %>%
  select(replicate) %>%
  mutate(area_with_qc = ifelse(str_detect(replicate, "Half"), 0.5, 1.0)) %>%
  mutate(metabolite = "Inj_vol")

is_stds_data <- qc_data %>%
  filter(metabolite %in% internal_standards$Compound_Name_Figure) %>%
  select(replicate, metabolite, area_with_qc) 

data_no_is <- qc_data %>%
  filter(!metabolite %in% internal_standards$Compound_Name_Figure)

is_stds_data <- rbind(is_stds_data, sampkey) %>%
  separate(replicate, into = c("a", "RunType", "c", "d"), sep = "_", remove = FALSE)
```

## plot internal standard areas
```{r}
is_stds_data %>% 
  # filter(RunType == "Poo") %>% 
  ggplot() + 
  geom_bar(aes(x = replicate, y = area_with_qc, color = RunType), stat = "identity", position = "dodge")  +
  facet_wrap( ~metabolite, scales = "free_y") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) +
  ggtitle("Internal Standard Raw Areas") 
```

## load data with no internal standards and get is conc and area data for BMIS
```{r}
data_long <- data_no_is %>%
  select(replicate, metabolite, area_with_qc) %>%
  filter(str_detect(replicate, "Poo")) %>%
  group_by(replicate) %>%
  rename(poo_area = area_with_qc) %>% 
  na.omit()

is_areas <- is_stds_data %>%
  filter(str_detect(replicate, "Poo")) %>%
  mutate(area_with_qc = ifelse(str_detect(replicate, "Half"), area_with_qc*2, area_with_qc)) %>% 
  filter(!str_detect(metabolite, "Inj_vol")) %>%
  select(-c(metabolite, a, RunType, c, d)) %>%
  group_by(replicate) %>%
  rename(int_area = area_with_qc) %>% 
  rename(internal_standard = metabolite) %>% 
  group_by(internal_standard) %>% 
  mutate(avg_is_area = mean(int_area))

is_areas_smp <- is_stds_data %>%
  filter(str_detect(replicate, "Smp")) %>%
  mutate(area_with_qc = ifelse(str_detect(replicate, "Half"), area_with_qc*2, area_with_qc)) %>% 
  filter(!str_detect(metabolite, "Inj_vol")) %>%
  select(-c(metabolite, a, RunType, c, d)) %>%
  group_by(replicate) %>%
  rename(int_area = area_with_qc) %>% 
  rename(internal_standard = metabolite) %>% 
  group_by(internal_standard) %>% 
  mutate(avg_is_area = mean(int_area))

is_conc <- internal_standards %>% 
  select(Compound_Name_Figure, Concentration_uM) %>% 
  rename(internal_standard = Compound_Name_Figure) %>% 
  rename(conc_um = Concentration_uM) %>% 
  mutate(internal_standard = str_remove(internal_standard, ", .*")) %>% 
  mutate(internal_standard = str_remove(internal_standard, "D"))

is_areas_inj_vol <- is_stds_data %>%
  filter(str_detect(replicate, "Poo")) %>%
  mutate(inj_vol = ifelse(str_detect(replicate, "Half"), 0.5, 1)) %>% 
  # filter(!str_detect(metabolite, "Inj_vol")) %>%
  select(-c(metabolite, a, RunType, c, d)) %>%
  group_by(replicate) %>% 
  rename(internal_standard = metabolite) %>% 
  group_by(internal_standard) %>% 
  filter(internal_standard != "Inj_vol")
```

## isotopologues 
```{r}
isologs <- data_no_is %>%
  select(replicate, metabolite, area_with_qc) %>% 
  left_join(is_areas_smp, by = c("replicate")) %>% 
  mutate(internal_standard = str_remove(internal_standard, ", .*")) %>% 
  mutate(internal_standard = str_remove(internal_standard, "D")) %>% 
  filter(metabolite %in% internal_standard) %>% 
  mutate(replicate = str_remove(replicate, "230616_"))
  # distinct(metabolite, .keep_all = TRUE)


isologs_quant <- isologs %>% 
  left_join(is_conc, by = c("internal_standard")) %>% 
  left_join(metadata_targ, by = c("replicate")) %>% 
  filter(metabolite == internal_standard) %>% 
 #  mutate(area_with_qc = area_with_qc / pc_um) %>% 
  mutate(umol_per_vial = (area_with_qc/int_area)*conc_um) %>% 
  mutate(umol_enviro = umol_per_vial*400*1e-6/vol_L_filt*1000) %>% 
  select(replicate, metabolite, treatment, area_with_qc, date, umol_enviro) %>% 
  mutate(nmol_enviro = umol_enviro*1000)
```

## visualization of isotopologue quantification
```{r}
colors <- c("C" = "plum2", "LH" = "lightcoral", "LL" = "gold", "RH" = "brown3", "RL" = "orange", "T0" = "grey50", 
                              "ZF" = "mediumpurple", "ZH" = "lightseagreen", "ZL" = "darkslategray2")

isologs_quant %>% 
  filter(metabolite == "Trimethylamine N-oxide") %>% 
  group_by(treatment, date, metabolite) %>% 
  mutate(avg_umol_enviro = mean(umol_enviro)) %>% 
  distinct(avg_umol_enviro, .keep_all = TRUE) %>% 
  ggplot() + 
  geom_line(aes(x = date, y = avg_umol_enviro, color = treatment)) + 
  scale_color_manual(values = colors) + 
  theme_bw()

isologs_quant %>% 
  filter(metabolite == "Trimethylamine N-oxide") %>% 
  group_by(treatment, date, metabolite) %>% 
  mutate(avg_umol_enviro = mean(umol_enviro)) %>% 
  distinct(avg_umol_enviro, .keep_all = TRUE) %>% 
  ggplot() + 
  geom_line(aes(x = date, y = avg_umol_enviro, color = treatment)) + 
  scale_color_manual(values = colors) + 
  theme_bw()
```



## RSDs of pooled
```{r}
rsd_data <- data_long %>%
  mutate(rsd_og = sd(poo_area)/mean(poo_area)) %>% 
  left_join(is_areas, by = c("replicate")) %>%
  filter(!metabolite %in% isologs$metabolite) %>%
  mutate(norm_area = poo_area/int_area*avg_is_area) %>%
  group_by(metabolite, internal_standard) %>%
  mutate(rsd = sd(norm_area)/mean(norm_area)) %>% 
  group_by(metabolite, internal_standard) %>%
  mutate(rsd_imp = (rsd_og - rsd)/rsd_og)

poo_bmis <- rsd_data %>%
  filter(rsd_imp > .4) %>%
  arrange(desc(rsd_imp)) %>%
  group_by(metabolite) %>%
  slice(1) %>%
  select(metabolite, internal_standard, rsd, int_area, avg_is_area) %>%
  rename(bmis = internal_standard) %>% 
  rename(rsd_poo = rsd)
```

## Sample BMIS
```{r}
sample_long <- data_no_is %>% 
  select(replicate, metabolite, num_area) %>% 
  filter(!str_detect(replicate, "Poo|Std")) %>% 
  filter(!metabolite %in% isologs$metabolite) %>% 
  left_join(poo_bmis) %>% 
  na.omit() %>% 
  mutate(norm_area = num_area/int_area*avg_is_area) %>%
  group_by(metabolite, bmis) %>%
  mutate(rsd = sd(norm_area)/mean(norm_area))

sample_long %>% 
  ggplot() +
  geom_point(aes(x = rsd_poo, y = rsd)) + 
  facet_wrap(~bmis)
```

# Quant
response factor = peak_area/concentration (use standards in water)
response factor ratio = rf(stdinmat) - rf(h20inmat) / rf(stdinh2o)
umol in vial = area / (rf * rf ratio)
umol in enviro = umol in vial * 400 uL * 10e^-6 / L filt * 1000 * dilution factor
```{r}
ingalls_stds_quant <- read_csv("https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv") %>% 
  select(metabolite_name = Compound_Name_Figure, metabolite = Compound_Name, conc_um = Concentration_uM, 
         empirical_formula = Empirical_Formula, z, ion_form = Ionization_Form, mix = HILIC_Mix) %>% 
  # filter(z == 1) %>% 
  na.omit()

filt_data_quant <- filt_data %>% 
  select(filename, vol_L_filt) %>% 
  rename(replicate = filename) %>% 
  mutate(replicate = str_remove(replicate, ".mzML"))

adj_bmis_data <- sample_long %>% 
  mutate(adj_peak_area = num_area/(int_area/avg_is_area)) %>% 
  full_join(peak_list) %>% 
  select(replicate, metabolite, bmis, adj_peak_area) %>% 
  mutate(replicate = str_remove(replicate, "230616_"))
 
rfs_mix1 <- peak_list %>% 
  mutate(replicate = str_remove(replicate, "230213_")) %>% 
  filter(str_detect(replicate, "Std")) %>% 
  select(replicate, metabolite, num_area) %>% 
  left_join(ingalls_stds_quant, by = c("metabolite")) %>% 
  na.omit() %>% 
  mutate(std_type = str_extract(replicate, "InH2O|InMatrix|H2OinMatrix_1|H2OinMatrix_2")) %>% 
  group_by(metabolite, std_type) %>% 
  filter(mix == "Mix1") %>% 
  # filter(std_type == "H2OinMatrix_1") %>% 
  filter(!str_detect(replicate, "Mix2")) %>% 
  filter(!str_detect(std_type, "H2OinMatrix_2")) %>% 
  # need to calculate RFs based on mix
  reframe(mean_area = mean(num_area, na.rm = TRUE), conc_um = unique(conc_um)) %>% 
  pivot_wider(names_from = std_type, values_from = mean_area) %>% 
  mutate(rf_h2o = InH2O/conc_um) %>% 
  mutate(rf_mat = (InMatrix - H2OinMatrix_1)/conc_um) %>% 
  mutate(rf_ratio = rf_mat / rf_h2o) %>% 
  na.omit() %>% 
  rename(H2OinMatrix = H2OinMatrix_1)
 
rfs_mix2 <- peak_list %>% 
  mutate(replicate = str_remove(replicate, "230213_")) %>% 
  filter(str_detect(replicate, "Std")) %>% 
  select(replicate, metabolite, num_area) %>% 
  left_join(ingalls_stds_quant, by = c("metabolite")) %>% 
  na.omit() %>% 
  mutate(std_type = str_extract(replicate, "InH2O|InMatrix|H2OinMatrix_1|H2OinMatrix_2")) %>% 
  group_by(metabolite, std_type) %>% 
  filter(mix == "Mix2") %>% 
  filter(!str_detect(replicate, "Mix1")) %>% 
  filter(!str_detect(std_type, "H2OinMatrix_1")) %>% 
  # need to calculate RFs based on mix
  reframe(mean_area = mean(num_area, na.rm = TRUE), conc_um = unique(conc_um)) %>% 
  pivot_wider(names_from = std_type, values_from = mean_area) %>% 
  mutate(rf_h2o = InH2O/conc_um) %>% 
  mutate(rf_mat = (InMatrix - H2OinMatrix_2)/conc_um) %>% 
  mutate(rf_ratio = rf_mat / rf_h2o) %>% 
  na.omit() %>% 
  rename(H2OinMatrix = H2OinMatrix_2)

rfs <- rfs_mix1 %>% 
  rbind(rfs_mix2)

quant_data <- peak_list %>% 
  mutate(replicate = str_remove(replicate, "230213_")) %>% 
  left_join(adj_bmis_data) %>% 
  mutate(real_area = ifelse(is.na(adj_peak_area), num_area, adj_peak_area)) %>% 
  select(metabolite, replicate, bmis, real_area) %>% 
  mutate(bmis = replace_na(bmis, "None")) %>% 
  left_join(rfs) %>% 
  select(metabolite, replicate, bmis, real_area, rf_h2o, rf_mat, rf_ratio) %>% 
  mutate(uM = real_area/rf_h2o * 0.0004/2 * 1/rf_ratio) %>% 
  mutate(nM = uM*1000) %>% 
  ## need to left_join with filt_data
  # mutate(umol_in_enviro = umol_in_vial * 400 * 10e-6/2) %>% 
  na.omit() %>% 
  filter(!metabolite %in% ingalls_standards$internal_standards) %>% 
  filter(metabolite != "DL-Histidine, 15N") %>% 
  filter(nM > -0.01) %>% 
  filter(!str_detect(replicate, "Std"))
  
quant_data %>%
  group_by(metabolite) %>%
  mutate(mean_nm=mean(nM)) %>%
  arrange(desc(mean_nm)) %>%
  ungroup() %>% 
  mutate(metabolite = fct_inorder(metabolite)) %>%
  ggplot(aes(y = metabolite, x = log10(nM))) +
  geom_boxplot() +
  theme_bw()
  
  
write.csv(quant_data, file = "../csvs/quant_data.csv", row.names = FALSE)
write.csv(metadataframe, file = "../csvs/metadataframe.csv", row.names = FALSE)
```

# Quantified Visualizations
```{r}
quant_data <- read_csv("csvs/quant_data.csv")

nb.cols <- 55
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)

data_vis <- quant_data %>% 
  mutate(treatment = str_extract(replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_extract(replicate, "27June|30June|14July|21July|27July")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "14July", "7-14-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "21July", "7-21-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "30June", "6-30-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "27July", "7-27-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "27June", "6-27-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%m-%d-%y")) %>% 
  select(metabolite, replicate, uM, nM, treatment, date) %>% 
  na.omit() %>% 
  filter(nM != "Inf")

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(nM >= 0.5) %>% 
  ggplot() + 
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "fill") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(nM >= 0.5) %>% 
  ggplot() + 
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(nM >= 0.5) %>% 
  ggplot() + 
  geom_col(aes(x = date, y = nmol_avg, fill = metabolite), position = "fill") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~treatment) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(nM >= 0.5) %>% 
  ggplot() + 
  geom_col(aes(x = date, y = nmol_avg, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~treatment) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()
``` 


# Stats
```{r}
set.seed(20)

quant_nmds <- data_vis %>% 
  arrange(desc(nM)) %>% 
  group_by(metabolite, replicate) %>% 
  slice(1) %>% 
  group_by(metabolite, treatment, date) %>% 
  summarise(avg_nmol = mean(nM)) %>% 
  ungroup() %>% 
  complete(metabolite, date, treatment) %>%
  mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment))
  # mutate(avg_nmol = as.character(avg_nmol))

quant_mat <- quant_nmds %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(avg_nmol)) %>%
  select(metabolite, norm_conc, id) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc") %>%
  column_to_rownames("id") %>%
  data.matrix()

mdsout <- quant_mat %>%
  metaMDS(k = 2, autotransform = FALSE)

mds_data <- metadataframe %>% 
  ungroup() %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  filter(str_detect(filename, "Smp")) %>% 
  mutate(treatment = str_remove(treatment, "\\d")) %>% 
  mutate(id = paste(date, treatment)) %>% 
  select(-c("replicate", "filename"))

mdsout$points %>%
  as.data.frame() %>%
  rownames_to_column("id") %>%
  left_join(mds_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, shape = factor(date)), size=4) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - Before Treatments", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) +
  theme_bw() + 
  ggtitle("Statistical Variation of Metabolite Concentration (nM) in PERI-DICE")


plot(mdsout)
```
add neg metabs

ratio of nutrients is more important on short term (1-2 weeks) whereas amount added (high dose) is more important on long term scale (3-4 weeks)

```{r}
quant_nmds_r <- data_vis %>% 
  filter(str_detect(replicate, "RH|RL|LH|LL")) %>% 
  arrange(desc(nmol_in_enviro)) %>% 
  group_by(metabolite, replicate) %>% 
  slice(1) %>% 
  group_by(metabolite, treatment, date) %>% 
  summarise(avg_nmol = mean(nmol_in_enviro)) %>% 
  ungroup() %>% 
  # complete(metabolite, date, treatment) %>% 
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>% 
  mutate(id = paste(date, treatment))
  # mutate(avg_nmol = as.character(avg_nmol))

quant_mat_r <- quant_nmds_r %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(avg_nmol)) %>%
  select(metabolite, norm_conc, id) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc") %>%
  column_to_rownames("id") %>%
  data.matrix()

mdsout_r <- quant_mat_r %>%
  metaMDS(k = 2, autotransform = FALSE)

mds_data_r <- metadataframe %>% 
  ungroup() %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  filter(str_detect(filename, "Smp")) %>% 
  mutate(treatment = str_remove(treatment, "\\d")) %>% 
  mutate(id = paste(date, treatment)) %>% 
  select(-c("replicate", "filename"))

mdsout_r$points %>%
  as.data.frame() %>%
  rownames_to_column("id") %>%
  left_join(mds_data_r) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, shape = factor(date)), size=4) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High Dose",
                                                                       "Low N:P, Low Dose",
                                                                       "Low N:P, High Dose",
                                                                        "Redfield Ratio - Low Dose",
                                                                       "")) +
  theme_bw() + 
  ggtitle("Statistical Variation of Metabolite Concentration (nM) in PERI-DICE")
```


# Volcano plots
```{r}

foldchange <- data.frame(metabolite = character(), date = character(), treatment = character(), pval = numeric(), fc = numeric())

data_vis_fc <- data_vis %>% 
  filter(!str_detect(replicate, "Tote"))

for (metabolite in unique(data_vis_fc$metabolite)) {
  for (group in unique(data_vis_fc$treatment)) {
    if (group != 'C') {
      temp <- data_vis_fc[data_vis_fc$metabolite == metabolite,]
      
      pval <- t.test(temp[temp$treatment == 'C',]$nmol_in_enviro,
                     temp[temp$treatment == group,]$nmol_in_enviro)$p.value
      fc <- (log2(mean(temp[temp$treatment == group,]$nmol_in_enviro) + 1) - 
               log2(mean(temp[temp$treatment == 'C',]$nmol_in_enviro) + 1))
      
      new_row <- data.frame(metabolite = metabolite, treatment = group, pval = pval, fc = fc)
      foldchange <- rbind(foldchange, new_row)
    }
  }
}


foldchange %>%
  mutate(pval = -log10(pval)) %>% 
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x = fc, y = pval, color = col_id)) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - No Added Nutrients", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Tote")) +
  theme_bw() +
  xlim(-6, 6) + 
  xlab("Log2(Fold-Change)") + 
  ylab("-Log10(p-value)") + 
  xlim(-2, 3) + 
  ggtitle("fold change calculated with respect to Control")
```

