---
title: "data_analysis_peridice"
author: "Iris Kern"
date: "2023-08-23"
output: html_document
---


```{r}
library(tidyverse)
library(readr) 
library(ggplot2)
library(vegan)
library(RColorBrewer)
library(plotly)
library(heatmaply)
```


```{r}
quant_data <- read_csv("csvs/quant_data.csv")
metadataframe <- read_csv("csvs/metadataframe.csv")
```

```{r}
data_an <- quant_data %>%  
  mutate(treatment = str_extract(replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_extract(replicate, "27June|30June|14July|21July|27July")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  select(metabolite, treatment, date, nM, replicate) %>% 
  filter(metabolite != "Sodium 2-mercaptoethanesulfonate") %>% 
  na.omit()


data_an %>%
  group_by(metabolite) %>%
  mutate(mean_nm=mean(nM)) %>%
  arrange(desc(mean_nm)) %>%
  ungroup() %>% 
  mutate(metabolite = fct_inorder(metabolite)) %>%
  ggplot(aes(y = metabolite, x = log10(nM))) +
  facet_wrap(~treatment + date) + 
  geom_boxplot() + 
  theme_bw()
```

# Quantified Visualizations
```{r}
set.seed(20)

nb.cols <- 170
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)

data_vis <- data_an

level_order <- c("Tote", "C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH")

peridice_fac <- data_vis %>% 
  mutate(treatment = as.factor(treatment))

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(nmol_avg < 0.5) %>% 
  filter(nmol_avg >= 1e-4) %>% 
  # filter(str_detect(treatment, "LL|LH|RL|RH")) %>% 
  filter(!str_detect(metabolite, "Fumaric acid")) %>% 
  ggplot() +
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "fill") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=5)) + 
  theme_bw()

# how to plot metabs where in at least 1 treatment/date their conc is > 0.5

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(!str_detect(metabolite, "Fumaric acid")) %>% 
  filter(str_detect(treatment, "Tote")) %>% 
  # filter(!str_detect(replicate, "Tote")) %>% 
  filter(nmol_avg >= 0.5) %>% 
  ggplot() + 
  geom_col(aes(x = date, y = nmol_avg, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~factor(treatment, levels = level_order)) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()
``` 


how to plot   
group_by(compound, treatment) %>%
  summarise(area=sum(area)) %>%
  arrange(desc(area)) %>% 
  group_by(treatment) %>% 
  do(head(., n = 10)) %>% 
  mutate(treatment, replace_na(treatment, "Tote")) %>% 
  rename(colid = `replace_na(treatment, "Tote")`)

```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(metabolite == "Dimethylsulfoniopropionate") %>% 
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() + 
  geom_line(aes(x = date, y = nmol_avg, color = col_id)) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - Before Treatments", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()

```

```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  mutate(sd = sd(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(metabolite == "Proline betaine") %>% 
  ggplot(aes(x = date, y = nmol_avg, color = treatment)) + 
  geom_line() + 
  geom_point() + 
  geom_errorbar(aes(ymin = nmol_avg - sd, ymax = nmol_avg + sd)) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()


data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  mutate(sd = sd(nM)) %>% 
  slice(1) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  # mutate(col_id=case_when(
  #   treatment=="C"~"plum2",
  #   treatment=="ZL"~"darkslategray2",
  #   treatment=="ZH"~"lightseagreen",
  #   treatment=="LL"~"gold",
  #   treatment=="LH"~"lightcoral",
  #   treatment=="RL"~"orange",
  #   treatment=="RH"~"brown3",
  #   treatment=="ZF"~"mediumpurple3", 
  #   TRUE~"grey50"
  # )) %>%
  filter(str_detect(metabolite, "Proline betaine")) %>% 
  # filter(str_detect(treatment, "RH|RL")) %>% 
  ggplot(aes(x = date, y = nmol_avg, color = factor(col_id, levels = level_order))) + 
  geom_line() + 
  geom_point() + 
  geom_errorbar(aes(ymin = nmol_avg - sd, ymax = nmol_avg + sd)) + 
  scale_color_identity(name = "treatment", guide = "legend", labels = c("Redfield Ratio - High Added N and P", "No N - Low Added P", 
                                                                        "Low N:P - Low Added N", "T0 - Before Treatments", 
                                                                        "Low N:P - High Added N", "No N - High Added P", "No Fe - High Added P", 
                                                                        "Redfield Ratio - Low Added N and P", "Control")) +
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw() + 
  # ggtitle("Timeseries") + 
  xlab("Date") + 
  ylab("nM (nmol per L)") 

level_order <- c("Tote", "C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH")
```
```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  mutate(sd = sd(nM)) %>% 
  slice(1) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  # mutate(col_id=case_when(
  #   treatment=="C"~"plum2",
  #   treatment=="ZL"~"darkslategray2",
  #   treatment=="ZH"~"lightseagreen",
  #   treatment=="LL"~"gold",
  #   treatment=="LH"~"lightcoral",
  #   treatment=="RL"~"orange",
  #   treatment=="RH"~"brown3",
  #   treatment=="ZF"~"mediumpurple3", 
  #   TRUE~"grey50"
  # )) %>%
  filter(str_detect(metabolite, "Proline betaine")) %>% 
  # filter(str_detect(treatment, "RH|RL")) %>% 
  ggplot(aes(x = date, y = nmol_avg, color = factor(treatment, levels = level_order))) + 
  geom_line() + 
  geom_point() + 
  # geom_errorbar(aes(ymin = nmol_avg - sd, ymax = nmol_avg + sd)) + 
  scale_color_manual(values = c("grey50", "plum2", "mediumpurple3", "darkslategray2",
                                "lightseagreen", "gold", "lightcoral", "orange", "brown3"), 
                     labels = c("T0", "Control - no nutrients added", "No Nitrogen or Phosphorus, low amount of Iron added", 
                                "No Nitrogen, low amount of Phosphorus and Iron added", "No Nitrogen, high amount of Phosphorus and Iron added", 
                                "Lower (than Redfield) ratio of N:P, low dose added", "Lower (than Redfield) ratio of N:P, high dose added", 
                                "Redfield Ratio of N:P, low dose added", "Redfield Ratio of N:P, high dose added"), 
                     name = "Treatment") + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw() + 
  ggtitle("Timeseries of Proline Betaine, Summarised by Triplicate") + 
  xlab("Date") + 
  ylab("nM (nmol per L)") 
```


```{r}


data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(nmol_avg >= 0.5) %>% 
  # filter(str_detect(treatment, "RH|RL|LH|LL")) %>% 
  filter(!str_detect(metabolite, "Fumaric acid")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  ggplot() +
  geom_col(aes(x = date, y = nmol_avg, fill = metabolite), position = "stack") +
  facet_wrap(~treatment) + 
  scale_fill_manual(values = mycolors) + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw()
```

```{r}
quant_nmds_heat <- data_vis %>% 
  # arrange(desc(nM)) %>% 
  # group_by(metabolite, replicate) %>% 
  # slice(1) %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nM)) %>%
  ungroup() %>%
  # complete(metabolite, date, treatment) %>%
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment))
  # filter(!str_detect(id, "Tote"))
  # mutate(replicate = str_remove(replicate, "230616_")) %>% 
  # rename(filename = replicate)
  # mutate(avg_nmol = as.character(avg_nmol))

heat_level_order_treat <- c("2022-06-27 Tote", "2022-06-30 C",  "2022-07-14 C",  "2022-07-21 C",  "2022-07-27 C",  
                            "2022-06-30 ZF", "2022-07-14 ZF", "2022-07-21 ZF", "2022-07-27 ZF",
                            "2022-06-30 ZL", "2022-07-14 ZL", "2022-07-21 ZL", "2022-07-27 ZL", 
                            "2022-06-30 ZH", "2022-07-14 ZH", "2022-07-21 ZH", "2022-07-27 ZH",
                            "2022-06-30 LL", "2022-07-14 LL", "2022-07-21 LL", "2022-07-27 LL",
                            "2022-06-30 LH", "2022-07-14 LH", "2022-07-21 LH", "2022-07-27 LH", 
                            "2022-06-30 RL", "2022-07-14 RL", "2022-07-21 RL", "2022-07-27 RL",
                            "2022-06-30 RH", "2022-07-14 RH", "2022-07-21 RH", "2022-07-27 RH")


heat_level_order_date <- c("2022-06-27 Tote", "2022-06-30 C","2022-06-30 ZF", "2022-06-30 ZL", 
                           "2022-06-30 ZH", "2022-06-30 LL","2022-06-30 LH", "2022-06-30 RL", "2022-06-30 RH",
                           "2022-07-14 C", "2022-07-14 ZF", "2022-07-14 ZL", "2022-07-14 ZH", "2022-07-14 LL", 
                           "2022-07-14 LH", "2022-07-14 RL", "2022-07-14 RH", 
                           "2022-07-21 C", "2022-07-21 ZF", "2022-07-21 ZL", "2022-07-21 ZH", 
                           "2022-07-21 LL", "2022-07-21 LH", "2022-07-21 RL", "2022-07-21 RH", 
                           "2022-07-27 C", "2022-07-27 ZF", "2022-07-27 ZL",  
                           "2022-07-27 ZH", "2022-07-27 LL", "2022-07-27 LH", "2022-07-27 RL",  "2022-07-27 RH")

quant_nmds_heat %>% 
  filter(str_detect(metabolite, "Cysteic acid"))

plotheatmap <- quant_nmds_heat %>% 
  group_by(metabolite) %>%
  mutate(norm_conc = rank(avg_nmol)) %>%
  # filter(str_detect(treatment, "RH|RL|C|Tote")) %>% 
  # filter(avg_nmol >= 0.01) %>% 
  select(metabolite, norm_conc, id) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
  mutate(id=factor(id, levels=c(heat_level_order_treat))) %>%
  arrange(id) %>%
  column_to_rownames("id") %>%
  data.matrix() %>% 
  heatmaply(fontsize_row = 10, fontsize_col = 7) # %>%
  # heatmaply(Rowv = FALSE, fontsize_row = 10, fontsize_col = 7)

plotheatmap
```


# Stats
```{r}
quant_nmds <- data_vis %>% 
  # arrange(desc(nM)) %>% 
  # group_by(metabolite, replicate) %>% 
  # slice(1) %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nM)) %>%
  ungroup() %>%
  # complete(metabolite, date, treatment) %>%
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment)) %>% 
  filter(!str_detect(id, "Tote"))
  # mutate(replicate = str_remove(replicate, "230616_")) %>% 
  # rename(filename = replicate)
  # mutate(avg_nmol = as.character(avg_nmol))

quant_mat <- quant_nmds %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(avg_nmol)) %>%
  select(metabolite, norm_conc, id) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
  column_to_rownames("id") %>%
  data.matrix() 

mdsout <- quant_mat %>%
  metaMDS(k = 2, autotransform = FALSE)

mds_data <- metadataframe %>% 
  ungroup() %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  filter(str_detect(filename, "Smp")) %>% 
  mutate(treatment = str_remove(treatment, "\\d")) %>% 
  mutate(id = paste(date, treatment))

# %>% 
#   select(-c("replicate", "filename"))

mdsout$points %>%
  as.data.frame() %>%
  rownames_to_column("id") %>%
  left_join(mds_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  mutate(date_fct = factor(date)) %>% 
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, shape = date_fct), size=4) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - Before Treatments", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) +
  theme_bw() + 
  ggtitle("Statistical Variation of Metabolite Concentration (nM) in PERI-DICE") + 
  facet_wrap(~date)
```

ratio of nutrients is more important on short term (1-2 weeks) whereas amount added (high dose) is more important on long term scale (3-4 weeks)

is there a way to look at metabolites that most affected (change conc most) in short term vs long term? which metabs are defined best by time vs treatment?


```{r}
quant_nmds_week <- data_vis %>% 
  # arrange(desc(nM)) %>% 
  # group_by(metabolite, replicate) %>% 
  # slice(1) %>% 
  # group_by(metabolite, treatment, date) %>% 
  # summarise(avg_nmol = mean(nM)) %>% 
  # ungroup() %>% 
  # complete(metabolite, date, treatment) %>%
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment)) %>% 
  mutate(replicate = str_remove(replicate, "230616_")) %>% 
  rename(filename = replicate) %>% 
  filter(str_detect(filename, "27July"))
  # mutate(avg_nmol = as.character(avg_nmol))

quant_mat_week <- quant_nmds_week %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(nM)) %>%
  select(metabolite, norm_conc, filename) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
  column_to_rownames("filename") %>%
  data.matrix() 

mdsout_week <- quant_mat_week %>%
  metaMDS(k = 2, autotransform = FALSE)


# %>% 
#   select(-c("replicate", "filename"))

mdsout_week$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(mds_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  mutate(date_fct = factor(date)) %>% 
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, shape = date_fct), size=4) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - Before Treatments", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) +
  theme_bw() + 
  ggtitle("Statistical Variation of Metabolite Concentration (nM) in PERI-DICE")
```


```{r}
quant_nmds_week_plotly <- data_vis %>% 
  # arrange(desc(nM)) %>% 
  # group_by(metabolite, replicate) %>% 
  # slice(1) %>% 
  # group_by(metabolite, treatment, date) %>%
  # summarise(avg_nmol = mean(nM)) %>%
  # ungroup() %>%
  # complete(metabolite, date, treatment) %>%
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment)) %>% 
  mutate(replicate = str_remove(replicate, "230616_")) %>% 
  rename(filename = replicate) 
  # filter(str_detect(filename, "Tote"))
  # mutate(avg_nmol = as.character(avg_nmol))


quant_mat_week_plotly <- quant_nmds_week_plotly %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(nM)) %>%
  select(metabolite, norm_conc, filename) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
  column_to_rownames("filename") %>%
  data.matrix() 

mdsout_week_plotly <- quant_mat_week_plotly %>%
  metaMDS(k = 3, autotransform = FALSE)

mds_data <- metadataframe %>% 
  ungroup() %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  filter(str_detect(filename, "Smp")) %>% 
  mutate(treatment = str_remove(treatment, "\\d")) %>% 
  mutate(id = paste(date, treatment))

data_vis %>% 
  filter(str_detect(metabolite, "Glucose"))

mdsout_week_plotly$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(mds_data) %>%
  # mutate(date_fct = factor(date)) %>% 
  # mutate(date_num = as.numeric(date_fct)) %>% 
  plot_ly(x= ~MDS1, y= ~MDS2, z = ~MDS3, color=~treatment) 


```

# Heatmap
```{r}
data_vis %>%filter(str_detect(metabolite, "Choline sulfate"))
```



fold change
calculate the fold change of metabolite nM from C at each timepoint
```{r}
foldchange <- data.frame(metabolite = character(), treatment = character(), pval = numeric(), fc = numeric())

for (metabolite in unique(data_vis$metabolite)) {
  for (group in unique(data_vis$treatment)) {
    if (treatment != 'Tote') {
      temp <- data_vis[data_vis$metabolite == metabolite,]
      # temp <- temp[temp$treatment == group,]
      
      pval <- t.test(temp[temp$treatment == 'Tote',]$nM,
                     temp[temp$treatment == treatment,]$nM)$p.value
      fc <- (log2(mean(temp[temp$treatment == treatment,]$nM) + 1) - 
               log2(mean(temp[temp$treatment == 'T0',]$nM) + 1))
      
      new_row <- data.frame(metabolite = metabolite, treatment = group, pval = pval, fc = fc)
      foldchange <- rbind(foldchange, new_row)
    }
  }
}
```




# Volcano plots
```{r}
foldchange <- data.frame(metabolite = character(), id = character(), pval = numeric(), fc = numeric())

data_vis_fc <- data_vis %>% 
  select(-c("replicate")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  pivot_wider(names_from = treatment, values_from = nM, values_fn = mean) %>% 
  pivot_longer(cols = c(ZF, ZL, ZH, LL, LH, RL, RH)) %>% 
  mutate(fold_change = log2(value/C))

data_vis_km <- data_vis %>% 
  pivot_wider(names_from = metabolite, values_from = nM) %>% 
  select(-c("treatment", "date")) %>% 
  column_to_rownames("replicate") %>% 
  data.matrix() %>%
  `[<-`(is.na(.), 0) %>%
  scale() %>%
  t() %>%
  kmeans(centers = 6) %>% 
  pluck("cluster") %>%
  as.data.frame() %>%
  set_names("cluster") %>%
  rownames_to_column("metabolite")

data_vis_cluster <- data_vis_fc %>% 
  left_join(data_vis_km)

data_vis_cluster %>% 
  ggplot(aes(x = date, y = fold_change, color = name, shape = factor(cluster))) +
  geom_point() + 
  geom_line()
```


```{r}
data_vis_fc %>% 
  na.omit() %>% 
  group_by(name, date) %>% 
  summarise(avg_fc = mean(fold_change)) %>% 
  ggplot(aes(x = date, y = avg_fc, color = name)) + 
  geom_point() + 
  geom_line()
  
```
# nutrient table
```{r}
treatment_table <- data.frame(col1 = c("C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH"), 
                              col2 = c("No nutrients added", "No N, No P, Low Fe", "No N, Low P, Low Fe", "No N, High P, Low Fe", 
                                       "6N:1P, Low Amount Added", "6N:1P, High Amount Added", 
                                       "16N:1P, Low Amount Added", "16N:1P, High Amount Added")) %>% 
  rename(Treatment = col1, Description = col2)

```



