---
title: "peridice_writeup"
author: "Iris Kern"
date: "2023-09-19"
output: html_document
---

# Setup
```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(plotly)

set.seed(20)
```

# Data
## Load in
```{r}
quant_data <- read_csv("csvs/quant_data.csv")
metadataframe <- read_csv("csvs/metadataframe.csv")
```

## Wrangle
```{r}
data_vis <- quant_data %>%  
  mutate(treatment = str_extract(replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_extract(replicate, "27June|30June|14July|21July|27July")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  select(metabolite, treatment, date, nM, replicate) %>% 
  filter(metabolite != "Sodium 2-mercaptoethanesulfonate") %>% 
  filter(metabolite != "Fumaric acid") %>% 
  na.omit()

level_order <- c("Tote", "C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH")

heat_level_order_treat <- c("2022-06-27 Tote", "2022-06-30 C",  "2022-07-14 C",  "2022-07-21 C",  "2022-07-27 C",  
                            "2022-06-30 ZF", "2022-07-14 ZF", "2022-07-21 ZF", "2022-07-27 ZF",
                            "2022-06-30 ZL", "2022-07-14 ZL", "2022-07-21 ZL", "2022-07-27 ZL", 
                            "2022-06-30 ZH", "2022-07-14 ZH", "2022-07-21 ZH", "2022-07-27 ZH",
                            "2022-06-30 LL", "2022-07-14 LL", "2022-07-21 LL", "2022-07-27 LL",
                            "2022-06-30 LH", "2022-07-14 LH", "2022-07-21 LH", "2022-07-27 LH", 
                            "2022-06-30 RL", "2022-07-14 RL", "2022-07-21 RL", "2022-07-27 RL",
                            "2022-06-30 RH", "2022-07-14 RH", "2022-07-21 RH", "2022-07-27 RH")

heat_level_order_date <- c("2022-06-27 Tote", "2022-06-30 C","2022-06-30 ZF", "2022-06-30 ZL", 
                           "2022-06-30 ZH", "2022-06-30 LL","2022-06-30 LH", "2022-06-30 RL", "2022-06-30 RH",
                           "2022-07-14 C", "2022-07-14 ZF", "2022-07-14 ZL", "2022-07-14 ZH", "2022-07-14 LL", 
                           "2022-07-14 LH", "2022-07-14 RL", "2022-07-14 RH", 
                           "2022-07-21 C", "2022-07-21 ZF", "2022-07-21 ZL", "2022-07-21 ZH", 
                           "2022-07-21 LL", "2022-07-21 LH", "2022-07-21 RL", "2022-07-21 RH", 
                           "2022-07-27 C", "2022-07-27 ZF", "2022-07-27 ZL",  
                           "2022-07-27 ZH", "2022-07-27 LL", "2022-07-27 LH", "2022-07-27 RL",  "2022-07-27 RH")
```

## Data vis setup
```{r}
nb.cols <- 170
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)
```

# Metabolome Overview
## Stacked barplots - relative concentrations
```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(treatment != "Tote") %>% 
  # filter(nmol_avg >= 0.5) %>% 
  ggplot() +
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "fill") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=3)) + 
  theme_bw() + 
  theme(legend.position = "none")
```

## Absolute metabolomic concentrations - biomass?
```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(treatment != "Tote") %>% 
  ggplot() +
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=5)) + 
  theme_bw() + 
  theme(legend.position = "none")
```

## Timeseries
```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  mutate(sd = sd(nM)) %>% 
  # slice(1) %>%
  distinct(nmol_avg, .keep_all = TRUE) %>%
  # mutate(col_id=case_when(
  #   treatment=="C"~"plum2",
  #   treatment=="ZL"~"darkslategray2",
  #   treatment=="ZH"~"lightseagreen",
  #   treatment=="LL"~"gold",
  #   treatment=="LH"~"lightcoral",
  #   treatment=="RL"~"orange",
  #   treatment=="RH"~"brown3",
  #   treatment=="ZF"~"mediumpurple3", 
  #   TRUE~"grey50"
  # )) %>%
  filter(str_detect(metabolite, "Proline betaine")) %>% 
  # filter(str_detect(treatment, "RH|RL")) %>% 
  ggplot(aes(x = date, y = nmol_avg, color = factor(treatment, levels = level_order))) +
  geom_line() +
  geom_point() +
  # geom_errorbar(aes(ymin = nmol_avg - sd, ymax = nmol_avg + sd)) + 
  scale_color_manual(values = c("grey50", "plum2", "mediumpurple3", "darkslategray2",
                                "lightseagreen", "gold", "lightcoral", "orange", "brown3"), 
                     labels = c("T0", "Control - no nutrients added", "No Nitrogen or Phosphorus, low amount of Iron added", 
                                "No Nitrogen, low amount of Phosphorus and Iron added", "No Nitrogen, high amount of Phosphorus and Iron added", 
                                "Lower (than Redfield) ratio of N:P, low dose added", "Lower (than Redfield) ratio of N:P, high dose added", 
                                "Redfield Ratio of N:P, low dose added", "Redfield Ratio of N:P, high dose added"), 
                     name = "Treatment") + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw() + 
  ggtitle("Timeseries of Proline Betaine, Summarised by Triplicate") + 
  xlab("Date") + 
  ylab("nM (nmol per L)") 
```

```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  mutate(sd = sd(nM)) %>% 
  # slice(1) %>%
  distinct(nmol_avg, .keep_all = TRUE) %>%
  # mutate(col_id=case_when(
  #   treatment=="C"~"plum2",
  #   treatment=="ZL"~"darkslategray2",
  #   treatment=="ZH"~"lightseagreen",
  #   treatment=="LL"~"gold",
  #   treatment=="LH"~"lightcoral",
  #   treatment=="RL"~"orange",
  #   treatment=="RH"~"brown3",
  #   treatment=="ZF"~"mediumpurple3", 
  #   TRUE~"grey50"
  # )) %>%
  filter(str_detect(metabolite, "Urea")) %>% 
  # filter(str_detect(treatment, "RH|RL")) %>% 
  ggplot(aes(x = date, y = nmol_avg, color = factor(treatment, levels = level_order))) +
  geom_line() +
  geom_point() +
  # geom_errorbar(aes(ymin = nmol_avg - sd, ymax = nmol_avg + sd)) + 
  scale_color_manual(values = c("grey50", "plum2", "mediumpurple3", "darkslategray2",
                                "lightseagreen", "gold", "lightcoral", "orange", "brown3"), 
                     labels = c("T0", "Control - no nutrients added", "No Nitrogen or Phosphorus, low amount of Iron added", 
                                "No Nitrogen, low amount of Phosphorus and Iron added", "No Nitrogen, high amount of Phosphorus and Iron added", 
                                "Lower (than Redfield) ratio of N:P, low dose added", "Lower (than Redfield) ratio of N:P, high dose added", 
                                "Redfield Ratio of N:P, low dose added", "Redfield Ratio of N:P, high dose added"), 
                     name = "Treatment") + 
  guides(fill=guide_legend(ncol=2)) + 
  theme_bw() + 
  ggtitle("Timeseries of Urea, Summarised by Triplicate") + 
  xlab("Date") + 
  ylab("nM (nmol per L)") 
```

# Fold change - specific metabolites
## calculating fold change and kmeans
```{r}
data_vis_fc <- data_vis %>% 
  select(-c("replicate")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  pivot_wider(names_from = treatment, values_from = nM, values_fn = mean) %>% 
  pivot_longer(cols = c(ZF, ZL, ZH, LL, LH, RL, RH)) %>% 
  mutate(fold_change = log2(value/C))

data_vis_km <- data_vis %>% 
  pivot_wider(names_from = metabolite, values_from = nM) %>% 
  select(-c("treatment", "date")) %>% 
  column_to_rownames("replicate") %>% 
  data.matrix() %>%
  `[<-`(is.na(.), 0) %>%
  scale() %>%
  t() %>%
  kmeans(centers = 9) %>% 
  pluck("cluster") %>%
  as.data.frame() %>%
  set_names("cluster") %>%
  rownames_to_column("metabolite")

data_vis_cluster <- data_vis_fc %>% 
  left_join(data_vis_km) %>% 
  rename(control = C, treatment = name, nM = value) %>% 
  mutate(cluster = as.numeric(cluster))
```

## visualizations
```{r}
data_vis_cluster %>% 
  na.omit() %>% 
  group_by(cluster, date, treatment) %>% 
  mutate(avg_fc = mean(fold_change)) %>% 
  ggplot(aes(x = date, y = avg_fc, color = factor(treatment, levels = level_order))) + 
  geom_point() +
  geom_line() + 
  theme_bw() + 
  facet_wrap(~cluster) +
  scale_color_manual(values = c("mediumpurple3", "darkslategray2",
                                "lightseagreen", "gold", "lightcoral", "orange", "brown3"),
                     labels = c("No Nitrogen or Phosphorus",
                                "No Nitrogen, low amount of Phosphorus", "No Nitrogen, high amount of Phosphorus",
                                "Lower (than Redfield) ratio of N:P, low dose added", "Lower (than Redfield) ratio of N:P, high dose added",
                                "Redfield Ratio of N:P, low dose added", "Redfield Ratio of N:P, high dose added"),
                     name = "Treatment")

```
The metabolites that are most influenced by ZF (cluster 1) are less influenced by nutrient ratios - indicates phosphorus limitation for those metabolites.
How do i visualize the metabolites in these categories and select a few?
```{r}
nb.cols <- 30
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)

plot <- data_vis_cluster %>% 
  na.omit() %>% 
  filter(cluster == 3) %>% 
  ggplot() +
  geom_col(aes(x = date, y = nM, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) +
  facet_wrap(~treatment)

ggplotly(plot)

plotline <- data_vis_cluster %>% 
  na.omit() %>% 
  filter(cluster == 4) %>% 
  ggplot() +
  geom_line(aes(x = date, y = fold_change, color = metabolite), position = "stack") +
  scale_color_manual(values = mycolors) +
  facet_wrap(~treatment)

ggplotly(plotline)

data_vis_cluster %>% 
  na.omit() %>% 
  filter(str_detect(metabolite, "Gonyol")) %>% 
  ggplot() +
  geom_line(aes(x = date, y = fold_change, color = metabolite), position = "stack") +
  # scale_color_manual(values = mycolors) +
  facet_wrap(~treatment)

```
5-oxoproline is associated with cyanobacteria, derivative of glutamic acid

gonyol and dimethysulfonioacetate - dinoflagellates and phytoplankton, could be potential alternatives than DMSP for marine sulfur cycle https://pubmed.ncbi.nlm.nih.gov/32113191/

5-methylcytosine - highest fold change in ZH, also ZF. Not high in ZL, why? Low P with iron is correct ratio? RL and ZL are only treatments where 5-methylcytosine declines
https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-022-01340-w 
https://www.frontiersin.org/articles/10.3389/fmicb.2019.01291/full 

N-acetyl-D-glucosamine, O-Propionylcarnitine, and Sarcosine all increase in RL first and then LL, no others, everything returns to base at end.

Melamine only increases in RL, but all return to base at end

Acetylcholine - peak at 2-week period and then peaks again at month in LL, RH, RL, ZH, ZL. Not in LH or ZF.

Acetylglycine - peaks at 2-week period and then peaks again at month in LH, RH, RL, ZH, ZL. Not in LL or ZF.

Chitobiose - associated with cyanobacteria and eukaryotic phytoplankton, upregulated in the LH, LL, RH, RL, ZF. Not in ZH or ZL.

Homarine - no trend really, de-regulated in RH at 2 week period

Sucrose - upregulated in the ZL, de-regulated in all others

Gonyol - upregulated in RL and LH

Key thoughts to explore: nutrient enrichment saturation? Interplay between ratios, P and N limiting each other?

2-ketoglutaric acid - key intermediate in the TCA cycle. Upregulated in RH, LH, higher in RL until very end. Lowest metabolite in all Z treatments. Disruption of the TCA cycle? LL has decreased until right at the very end of the month. 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4651028/ 

DMSP : https://www.nature.com/articles/s41396-023-01375-3 

Phytoplankton: https://journals.asm.org/doi/full/10.1128/msystems.01334-20#fig2 

MUST READ: https://www.frontiersin.org/articles/10.3389/fmars.2022.864796/full DONE
https://journals.asm.org/doi/full/10.1128/spectrum.04000-22 
https://www.dl.begellhouse.com/journals/7dd4467e7de5b7ef,3992b3431844bd87,35aa504b60398141.html 


OVERALL:
Which metabolites are upregulated in only Rs? Only Ls? Only Zs? are there metabolites that are only de-regulated in the Zs? are there metabolites that are de-regulated in the Ls? Rs? What about the relationship between L and H? How does ZF compare to ZL/H?

What do these trends indicate of the microbial community health and composition? Does rate or ratio have more of an impact on the metabolic activity of the community? On a 2-week scale do the metabolic trends differ from the final 2 weeks? Maybe split up by week?


On a 2 week scale:
```{r}
biweek_clust <- data_vis_cluster %>% 
  na.omit() %>% 
  group_by(cluster, date, treatment) %>% 
  mutate(avg_fc = mean(fold_change)) %>% 
  filter(str_detect(date, "7-14|7-21|7-27")) %>% 
  ggplot(aes(x = date, y = avg_fc, color = factor(treatment, levels = level_order))) + 
  geom_point() +
  geom_line() + 
  theme_bw() + 
  facet_wrap(~cluster) +
  scale_color_manual(values = c("mediumpurple3", "darkslategray2",
                                "lightseagreen", "gold", "lightcoral", "orange", "brown3"),
                     labels = c("No Nitrogen or Phosphorus",
                                "No Nitrogen, low amount of Phosphorus", "No Nitrogen, high amount of Phosphorus",
                                "Lower (than Redfield) ratio of N:P, low dose added", "Lower (than Redfield) ratio of N:P, high dose added",
                                "Redfield Ratio of N:P, low dose added", "Redfield Ratio of N:P, high dose added"),
                     name = "Treatment")

ggplotly(biweek_clust)
```
Different groups: (+) {-}
1: (LH > RH > ZH > RL > LL > ZL) > {ZF}
LH is a lot higher foldchange, RH, ZH, RL all group together. LL and ZL group together.

2: (RH > LL > RL > LH > ZH) > {ZL > ZF}
RH is a lot higher foldchange, LL, RL, LH, ZH all group together, ZL and ZF group together.

3: (RH > RL > ZL > ZH > LL > LH) > {ZF}
RH, RL, ZL, ZH all group together, LL and LH group, ZF on the bottom.

4. (RH) > {LH > RL > LL > ZH > ZL > ZF}
All very similar, all start below 0 and have positive trend

5. (ZF > LL > LH > ZL) > {RL > ZL > ZH} 
All group together, all other than ZF start above 0. 

6. (RL > RH > LL > LH > ZH > ZL) > {ZF} 


7. (LL > LH > ZF > RL > ZL > ZH > RH) 
8. (RH > RL > LH) > {LL > ZL > ZH > ZF}
9. (RH > RL > LH) > {ZF > ZH > LL > ZL }
