---
title: "peridice_summary"
author: "Iris Kern"
date: "2023-10-02"
output: html_document
---

# FIGURES FOR SCOPE AND SECOND YEAR TALK
heatmap (clustered), NMDS (RH/RL/LH/LL), metab group 1 and group 2 - kmeans clusters

# Setup
```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(vegan)
library(heatmaply)

set.seed(20)
```

# Data
## Load in
```{r}
quant_data <- read_csv("csvs/quant_data.csv")
metadataframe <- read_csv("csvs/metadataframe.csv")
```

## Wrangle
```{r}
data_vis <- quant_data %>%  
  mutate(treatment = str_extract(replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_extract(replicate, "27June|30June|14July|21July|27July")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  select(metabolite, treatment, date, nM, replicate) %>% 
  filter(metabolite != "Sodium 2-mercaptoethanesulfonate") %>% 
  filter(metabolite != "Fumaric acid") %>% 
  na.omit()

level_order <- c("Tote", "C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH")

heat_level_order_treat <- c("2022-06-27 Tote", "2022-06-30 C",  "2022-07-14 C",  "2022-07-21 C",  "2022-07-27 C",  
                            "2022-06-30 ZF", "2022-07-14 ZF", "2022-07-21 ZF", "2022-07-27 ZF",
                            "2022-06-30 ZL", "2022-07-14 ZL", "2022-07-21 ZL", "2022-07-27 ZL", 
                            "2022-06-30 ZH", "2022-07-14 ZH", "2022-07-21 ZH", "2022-07-27 ZH",
                            "2022-06-30 LL", "2022-07-14 LL", "2022-07-21 LL", "2022-07-27 LL",
                            "2022-06-30 LH", "2022-07-14 LH", "2022-07-21 LH", "2022-07-27 LH", 
                            "2022-06-30 RL", "2022-07-14 RL", "2022-07-21 RL", "2022-07-27 RL",
                            "2022-06-30 RH", "2022-07-14 RH", "2022-07-21 RH", "2022-07-27 RH")

heat_level_order_date <- c("2022-06-27 Tote", "2022-06-30 C","2022-06-30 ZF", "2022-06-30 ZL", 
                           "2022-06-30 ZH", "2022-06-30 LL","2022-06-30 LH", "2022-06-30 RL", "2022-06-30 RH",
                           "2022-07-14 C", "2022-07-14 ZF", "2022-07-14 ZL", "2022-07-14 ZH", "2022-07-14 LL", 
                           "2022-07-14 LH", "2022-07-14 RL", "2022-07-14 RH", 
                           "2022-07-21 C", "2022-07-21 ZF", "2022-07-21 ZL", "2022-07-21 ZH", 
                           "2022-07-21 LL", "2022-07-21 LH", "2022-07-21 RL", "2022-07-21 RH", 
                           "2022-07-27 C", "2022-07-27 ZF", "2022-07-27 ZL",  
                           "2022-07-27 ZH", "2022-07-27 LL", "2022-07-27 LH", "2022-07-27 RL",  "2022-07-27 RH")
```

# heatmap
```{r}
quant_nmds_heat <- data_vis %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nM)) %>%
  ungroup() %>%
  mutate(id = paste(date, treatment)) %>% 
  filter(str_detect(treatment, "RH|RL|LH|LL"))

plotheatmap <- quant_nmds_heat %>% 
  # group_by(metabolite) %>%
  # filter(str_detect(treatment, "RH|RL|C|Tote")) %>% 
  # filter(avg_nmol >= 0.01) %>% 
  complete(metabolite, id, fill = list(avg_nmol = 0)) %>% 
  group_by(metabolite) %>% 
  mutate(norm_conc = rank(avg_nmol)) %>%
  select(metabolite, norm_conc, id) %>%
  pivot_wider(names_from = "metabolite", values_from = "norm_conc") %>%
  mutate(id=factor(id, levels=c(heat_level_order_treat))) %>%
  arrange(id) %>%
  column_to_rownames("id") %>%
  data.matrix() %>% 
  heatmaply(fontsize_row = 10, fontsize_col = 7) # %>%
  # heatmaply(Rowv = FALSE, fontsize_row = 10, fontsize_col = 7)

# make clustered heatmap from scratch using pheatmap

plotheatmap
```


## Data vis setup
```{r}
nb.cols <- 170
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)
```

# Metabolome Overview
## Stacked barplots - relative concentrations
```{r}
data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(treatment != "Tote") %>% 
  # filter(nmol_avg >= 0.5) %>% 
  ggplot() +
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "fill") +
  # scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=3)) + 
  theme_bw() + 
  theme(legend.position = "none")
```


```{r}
nb.cols <- 170
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)

data_vis %>% 
  group_by(metabolite, treatment, date) %>% 
  mutate(nmol_avg = mean(nM)) %>% 
  distinct(nmol_avg, .keep_all = TRUE) %>% 
  filter(!str_detect(treatment, "Tote|ZF|ZL|ZH")) %>% 
  ggplot() +
  geom_col(aes(x = treatment, y = nmol_avg, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~date) + 
  guides(fill=guide_legend(ncol=5)) + 
  theme_bw() + 
  theme(legend.position = "none")
```

# Fold change - specific metabolites
## calculating fold change and kmeans
```{r}
data_vis_fc <- data_vis %>% 
  select(-c("replicate")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  pivot_wider(names_from = treatment, values_from = nM, values_fn = mean) %>% 
  pivot_longer(cols = c(ZF, ZL, ZH, LL, LH, RL, RH)) %>% 
  mutate(fold_change = log2(value/C))

data_vis_km <- data_vis %>% 
  pivot_wider(names_from = metabolite, values_from = nM) %>% 
  select(-c("treatment", "date")) %>% 
  column_to_rownames("replicate") %>% 
  data.matrix() %>%
  `[<-`(is.na(.), 0) %>%
  scale() %>%
  t() %>%
  kmeans(centers = 6) %>% 
  pluck("cluster") %>%
  as.data.frame() %>%
  set_names("cluster") %>%
  rownames_to_column("metabolite")

data_vis_cluster <- data_vis_fc %>% 
  left_join(data_vis_km) %>% 
  rename(control = C, treatment = name, nM = value) %>% 
  mutate(cluster = as.numeric(cluster))
```


```{r}
data_vis_cluster %>% 
  na.omit() %>% 
  group_by(cluster, date, treatment) %>% 
  filter(str_detect(treatment, "RH|RL|LH|LL")) %>% 
  mutate(avg_fc = mean(fold_change)) %>% 
  ggplot(aes(x = date, y = avg_fc, color = factor(treatment, levels = level_order))) + 
  geom_point() +
  geom_line() + 
  theme_bw() + 
  facet_wrap(~cluster) +
  scale_color_manual(values = c("mediumpurple3", "darkslategray2",
                                "lightseagreen", "gold", "lightcoral", "orange", "brown3"),
                     labels = c("No Nitrogen or Phosphorus",
                                "No Nitrogen, low amount of Phosphorus", "No Nitrogen, high amount of Phosphorus",
                                "Lower (than Redfield) ratio of N:P, low dose added", "Lower (than Redfield) ratio of N:P, high dose added",
                                "Redfield Ratio of N:P, low dose added", "Redfield Ratio of N:P, high dose added"),
                     name = "Treatment") +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5))
```
Summary of 6 different groups:
Group 1: no change, unaffected by nutrient changes
Group 2: LH,LL increased at 2/3 week level. all others no change (ZL decreased foldchange)
Group 3: spike in all treatments apart from ZH, all decrease at 3 week, then move back up with ZH increasing the most, second is ZL.
Group 4: spike in RH, LH, none others.
Group 5: Spike in RH, RL at 2 week mark, LH spikes at 3. All return to similar range at 4 week mark.
Group 6: RH, RL steady increase across 4 weeks, same with LH but that dips down at 4 week mark. All others decrease at 2 weeks, come back up at 4.

Why are different metabolites responding in these different ways?

```{r}
plot <- data_vis_cluster %>% 
  na.omit() %>% 
  filter(cluster == 1) %>% 
  ggplot() +
  geom_col(aes(x = date, y = nM, fill = metabolite), position = "stack") +
  scale_fill_manual(values = mycolors) +
  facet_wrap(~treatment)

ggplotly(plot)

data_clust <- data_vis_cluster %>%  
  select(cluster, metabolite) %>% 
  distinct(metabolite, .keep_all = TRUE)
```

Is it rate? or is it ratio? Interplay between RH,RL,LH,LL. 
ANOVA/NMDS and k-means
```{r}

```

