---
title: "peridice_comp"
author: "Iris Kern"
date: "2023-12-14"
output: html_document
---

# Setup
## Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(vegan)
library(heatmaply)
library(ggnewscale)
library(readr)
library(readxl)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(stats)
library(ggfortify)
library(ggh4x)

set.seed(20)
```

data setup
general exploration (heatmaps, nmds, split by metabolite function, pca, metabolite conc abs and rel - is this norm to pc?) 
narrow in on what comes out of that, stories
fun things (machine learning! maybe rfr? predicting pc from metab conc, dissolved related to par?)

Experiment intro

# Data Setup
```{r}
quant_data <- read_csv("csvs/quant_data.csv") %>% 
  filter(nM > 0) %>% 
  filter(nM < 100) 
  
metadataframe <- read_csv("csvs/metadataframe.csv")
pcpn <- read_xlsx("metadata/peridice_pcpn.xlsx") %>% 
  select(tank = Tank, treatment = Treatment2, date = Date, pn = `PN (uM)`, pc = `PC (uM)`, cn = Cnratio,added_N_uM = AddN) %>% 
  mutate(triplicate = str_extract(tank, "\\d")) %>% 
  mutate(tank = str_remove (tank, "1/2|1/3|2/2"))  %>% 
  group_by(tank, date) %>% 
  mutate(mean_pc = mean(pc)) %>% 
  distinct(mean_pc, .keep_all = TRUE) %>% 
  mutate(replicate = paste0("230616_Smp_", ""))
  

data_vis <- quant_data %>%  
  mutate(treatment = str_extract(replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote1|Tote2|Tote3|Tote4")) %>% 
  mutate(timepoint = str_extract(replicate, "27June|30June|14July|21July|27July")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  select(metabolite, treatment, date, nM, replicate) %>% 
  filter(metabolite != "Sodium 2-mercaptoethanesulfonate") %>% 
  filter(metabolite != "Fumaric acid") %>% 
  na.omit() %>% 
  mutate(triplicate = str_extract(replicate, "\\d$")) 

level_order <- c("T0", "C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH")

heat_level_order_treat <- c("2022-06-27 Tote", "2022-06-30 C",  "2022-07-14 C",  "2022-07-21 C",  "2022-07-27 C",  
                            "2022-06-30 ZF", "2022-07-14 ZF", "2022-07-21 ZF", "2022-07-27 ZF",
                            "2022-06-30 ZL", "2022-07-14 ZL", "2022-07-21 ZL", "2022-07-27 ZL", 
                            "2022-06-30 ZH", "2022-07-14 ZH", "2022-07-21 ZH", "2022-07-27 ZH",
                            "2022-06-30 LL", "2022-07-14  LL", "2022-07-21 LL", "2022-07-27 LL",
                            "2022-06-30 LH", "2022-07-14 LH", "2022-07-21 LH", "2022-07-27 LH", 
                            "2022-06-30 RL", "2022-07-14 RL", "2022-07-21 RL", "2022-07-27 RL",
                            "2022-06-30 RH", "2022-07-14 RH", "2022-07-21 RH", "2022-07-27 RH")

heat_level_order_date <- c("2022-06-27 Tote", "2022-06-30 C","2022-06-30 ZF", "2022-06-30 ZL", 
                           "2022-06-30 ZH", "2022-06-30 LL","2022-06-30 LH", "2022-06-30 RL", "2022-06-30 RH",
                           "2022-07-14 C", "2022-07-14 ZF", "2022-07-14 ZL", "2022-07-14 ZH", "2022-07-14 LL", 
                           "2022-07-14 LH", "2022-07-14 RL", "2022-07-14 RH", 
                           "2022-07-21 C", "2022-07-21 ZF", "2022-07-21 ZL", "2022-07-21 ZH", 
                           "2022-07-21 LL", "2022-07-21 LH", "2022-07-21 RL", "2022-07-21 RH", 
                           "2022-07-27 C", "2022-07-27 ZF", "2022-07-27 ZL",  
                           "2022-07-27 ZH", "2022-07-27 LL", "2022-07-27 LH", "2022-07-27 RL",  "2022-07-27 RH")

data_vis_norm <- data_vis %>% 
  group_by(replicate, date, metabolite) %>% 
  mutate(nM = mean(nM)) %>% 
  left_join(pcpn, by = c("treatment", "date", "triplicate")) %>% 
  mutate(nmol_per_pc = nM/pc) %>% 
  select(replicate = replicate.x, metabolite, treatment, date, nM, triplicate, nmol_per_pc) %>% 
  mutate(treatment = str_remove(treatment, "\\d"))

metab_groups <- read.csv(
  "https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv",
                              stringsAsFactors = FALSE, header = TRUE) %>% 
  select(metabolite = Compound_Name, emp_form = Empirical_Formula, metab_type = Compound_Type)

```

## Quality Control
```{r}
metab_qual <- read_csv("csvs/peak_quality.csv") %>% 
  select(metabolite = Metabolite, quality = Quality)

data_vis_qual <- data_vis_norm %>% 
  left_join(metab_qual, by = c("metabolite")) %>% 
  filter(quality == "Good") %>% 
  mutate(ratio = str_extract(treatment, "R|L|Z|C|Tote")) %>% 
  mutate(rate = str_extract(treatment,  "H$|L$|C|F|Tote")) %>% 
  left_join(metab_groups, by = c("metabolite")) %>% 
  distinct() %>% 
  filter(metab_type != "Cursed")
```

## Metadata
```{r}
pcpn %>% 
  mutate(cn_rank=(cn-min(cn))/(max(cn)-min(cn))) %>%
  filter(str_detect(treatment, "RH|RL|LH|LL|C")) %>% 
  group_by(triplicate, treatment, date) %>%  
  # mutate(avg_cn = mean(cn)) %>% 
  # filter(triplicate == 2) %>% 
  ggplot(aes(x = date, y = cn, group = interaction(treatment, triplicate), color = treatment)) +
  geom_line() + 
  geom_point() + 
  # scale_color_manual(values = c("plum2", "gold", "lightcoral", "darkorange", "brown3"), name = "Treatment") +
  facet_wrap(~treatment) + 
  theme_bw() +  
  ylab("C to N Ratio") + 
  xlab("Date")

pcpn %>% 
  mutate(pc_rank=(pc-min(pc))/(max(pc)-min(pc))) %>%
  filter(!str_detect(treatment, "Tote")) %>% 
  group_by(triplicate, treatment, date) %>%  
  # mutate(avg_cn = mean(cn)) %>% 
  # filter(triplicate == 2) %>% 
  ggplot(aes(x = date, y = pc, group = interaction(treatment, triplicate), color = treatment)) +
  geom_line() + 
  geom_point() + 
  # scale_color_manual(values = c("plum2", "gold", "lightcoral", "darkorange", "brown3"), name = "Treatment") +
  facet_grid(~treatment) + 
  theme_bw() +  
  ylab("Particulate Carbon") + 
  xlab("Date")
```
Particulate Carbon Trends:
C - no trend (slight spike at 2 week)
LH - first spike at 2 weeks, down at 3, way higher at 4
LL - peak at 2 weeks, down after
RH - sharp increase all 4 weeks
RL - peak at 2 weeks, down at 3, up at 4
ZF - small growth up 4 weeks
ZH - 2 week spike, down at 3, up higher at 4
ZL - max peak at 2 weeks, down at 3, slightly up at 4.

# Stats
## NMDS
```{r}
quant_nmds <- data_vis_qual %>% 
  # arrange(desc(nM)) %>% 
  # group_by(metabolite, replicate) %>% 
  # slice(1) %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nmol_per_pc)) %>%
  ungroup() %>%
  # complete(metabolite, date, treatment) %>%
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment)) %>% 
  filter(!str_detect(treatment, "Tote"))

quant_mat <- quant_nmds %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(avg_nmol)) %>%
  select(metabolite, norm_conc, id) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
  column_to_rownames("id") %>%
  data.matrix() 

mdsout <- quant_mat %>%
  metaMDS(k = 2, autotransform = FALSE)

mds_data <- metadataframe %>% 
  ungroup() %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  filter(str_detect(filename, "Smp")) %>% 
  mutate(treatment = str_remove(treatment, "\\d")) %>% 
  mutate(id = paste(date, treatment))


mdsout$points %>%
  as.data.frame() %>%
  rownames_to_column("id") %>%
  left_join(mds_data) %>%
  mutate(date_fct = factor(date)) %>% 
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=factor(treatment, levels = level_order), shape = date_fct), size=4) +
  scale_color_manual(values = c("plum2","mediumpurple3","darkslategray2", "lightseagreen","gold", "lightcoral","orange", "brown3"),
                     labels = c("Control","0N:1P:1Fe, high dose P","0N:1P, low dose added", "0N:1P, high dose added",
                                "6N:1P, low dose added", "6N:1P, high dose added",
                                "16N:1P, low dose added", "16N:1P, high dose added"),
                     name = "Treatment") +
  scale_shape_discrete(name = "Date") + 
  theme_bw() + 
  ggtitle("Variation of Metabolite Concentration (nM) Across Treatments")
```
Key takeaways: temporal shift but when normalizing to pc, the trends in treatments are more distinct. The controls as well as the 0N treatments remain closer to the 6-30 treatments, but the 6N,16N treatments branch out more.

Cool thoughts: maybe make nmds for ratio/rate? that's normalized to pc?

```{r}
# quant_nmds_rate <- data_vis_qual %>%
#   group_by(metabolite, rate, date) %>%
#   summarise(avg_nmol = mean(nmol_per_pc)) %>%
#   ungroup() %>% 
#   mutate(id = paste(date, rate)) %>%
#   filter(!str_detect(rate, "Tote"))
# 
# quant_mat_rate <- quant_nmds_rate %>%
#   group_by(metabolite) %>%
#   mutate(norm_conc = rank(avg_nmol)) %>%
#   select(metabolite, norm_conc, id) %>%
#   pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
#   column_to_rownames("id") %>%
#   data.matrix()
# 
# mdsout_rate <- quant_mat_rate %>%
#   metaMDS(k = 2, autotransform = FALSE)
# 
# mds_data_rate <- metadataframe %>%
#   ungroup() %>%
#   mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>%
#   mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>%
#   mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>%
#   filter(str_detect(filename, "Smp")) %>%
#   mutate(rate = str_extract(treatment,  "H$|L$|C|F|Tote")) %>%
#   mutate(id = paste(date, rate))
# 
# 
# mdsout_rate$points %>%
#   as.data.frame() %>%
#   rownames_to_column("id") %>%
#   left_join(mds_data) %>%
#   mutate(date_fct = factor(date)) %>%
#   ggplot() +
#   geom_point(aes(x=MDS1, y=MDS2, shape = date_fct), size=4) +
#   # scale_color_manual(values = c("plum2","mediumpurple3","darkslategray2", "lightseagreen","gold", "lightcoral","orange", "brown3"),
#   #                    labels = c("Control","0N:1P:1Fe, high dose P","0N:1P, low dose added", "0N:1P, high dose added",
#   #                               "6N:1P, low dose added", "6N:1P, high dose added",
#   #                               "16N:1P, low dose added", "16N:1P, high dose added"),
#   #                    name = "Treatment") +
#   scale_shape_discrete(name = "Date") +
#   theme_bw() +
#   ggtitle("Variation of Metabolite Concentration (nM) Across Treatments")
```


### PCA
generate pca and check variations explained by each principal component
```{r}
rate_ratio <- data_vis_qual %>% 
  mutate(ratio = str_extract(treatment, "R|L|Z|C|Tote")) %>% 
  mutate(rate = str_extract(treatment,  "H$|L$|C|F|Tote")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  # mutate(rate = str_replace(rate, "F", "H")) %>%
  mutate(replicate = str_remove(replicate, "230616_Smp_"))

pca_mat <- rate_ratio %>% 
  select(metabolite, replicate, nmol_per_pc) %>% 
  group_by(metabolite) %>% 
  # mutate(nmol_per_pc = (avg_nM - min(avg_nmol)) / (max(avg_nmol) - min(avg_nmol))) %>% 
  pivot_wider(names_from = "metabolite", values_from = "nmol_per_pc") %>% 
  column_to_rownames("replicate") %>% 
  data.matrix() %>% 
  `[<-`(is.na(.), 0)

prcomp(pca_mat) %>%
  pluck("x") %>%
  as.data.frame() %>%
  select(PC1:PC5) %>%
  rownames_to_column("replicate") %>%
  pivot_longer(starts_with("PC"), names_to = "PC", values_to = "PC_val") %>%
  left_join(rate_ratio) %>%
  ggplot() +
  geom_col(aes(x=replicate, y=PC_val, fill=ratio)) +
  # scale_fill_manual(values = c("plum2","gold", "brown3", "lightseagreen"),
  #                    labels = c("Control","6N:1P", "16N:1P", "0N:1P"),
  #                    name = "Treatment") + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5)) + 
  facet_wrap(~PC, ncol=1)

peri_pca_analysis <- prcomp(pca_mat) %>%
  pluck("x") %>%
  as.data.frame() %>%
  select(PC1:PC2) %>%
  rownames_to_column("replicate") %>%
  # pivot_longer(starts_with("PC"), names_to = "PC", values_to = "PC_val") %>%
  left_join(rate_ratio, by = "replicate")  %>% 
  mutate(PC1 = as.numeric(PC1)) %>% 
  mutate(PC2 = as.numeric(PC2))

```

plot pca
```{r}
peri_pca_analysis %>% 
  mutate(ratio = str_replace(ratio, "R", "16N:1P")) %>% 
  mutate(ratio = str_replace(ratio, "L", "6N:1P")) %>%
  mutate(ratio = str_replace(ratio, "Z", "0N:1P")) %>%
  # filter(str_detect(replicate, "21July")) %>% 
  filter(str_detect(rate, "H|L")) %>% 
  mutate(rate = str_replace(rate, "H", "High Dose of Nutrients")) %>% 
  mutate(rate = str_replace(rate, "L", "Low Dose of Nutrients")) %>% 
  ggplot(aes(x = PC1, y = PC2, group = factor(ratio, levels = c("0N:1P", "6N:1P", "16N:1P")))) + 
  # geom_point(aes(fill = ratio, color = ratio), size = 0.5) + 
  geom_point(aes(color = ratio), size = 0.75) + 
  geom_polygon(aes(fill = ratio), alpha = 0.5) + 
  #scale_color_discrete(name = "Ratio") + 
  theme_bw() + 
  facet_grid2(~rate~date, scales = "free", independent = "y")  + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, size = 10), axis.text.y = element_text(size = 10))

peri_pca_analysis %>% 
  mutate(rate = str_replace(rate, "H", "High Dose of Nutrients")) %>% 
  mutate(rate = str_replace(rate, "L", "Low Dose of Nutrients")) %>% 
  mutate(rate = str_replace(rate, "F", "High Dose of Nutrients-no Fe")) %>% 
  # filter(str_detect(replicate, "21July")) %>% 
  filter(str_detect(ratio, "R|L|Z")) %>% 
  # filter(str_detect(rate, "H|L")) %>% 
  mutate(ratio = str_replace(ratio, "R", "16N:1P")) %>% 
  mutate(ratio = str_replace(ratio, "L", "6N:1P")) %>%
  mutate(ratio = str_replace(ratio, "Z", "0N:1P")) %>%
  ggplot(aes(x = PC1, y = PC2, group = rate)) + 
  geom_point(aes(color = rate), size = 0.75) + 
  geom_polygon(aes(fill = rate), alpha = 0.5) + 
  # scale_fill_continuous(values = mycolors) + 
  theme_bw() + 
  facet_grid2(~factor(ratio, levels = c("0N:1P", "6N:1P", "16N:1P"))~date, scales = "free", independent = "y")  + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, size = 7), axis.text.y = element_text(size = 7))
```
Key takeaways: high dose nutrients split up by ratio - become more selective? Low dose nutrients group 16N and 6N together (week 2, nowhere else). Everything shifts at the 4 week mark, metabolome has shifted but what is that shift? More AAs? 

nutrients with 0N never split by rate of nutrient addition. 6N and 16N split by nutrient rate but only really on week 3. This is also again showing a divergence at 4 week mark - shift of metabolome, what is it?

### Heatmap
Grouping - metabolites present at beginning (T0) versus middle (7/14-7/21) versus end (7/27)
```{r}
quant_nmds_heat <- data_vis_qual %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nmol_per_pc)) %>%
  ungroup() %>%
  mutate(id = paste(date, treatment)) 

heatmap_data <- quant_nmds_heat %>% 
  complete(metabolite, id, fill = list(avg_nmol = 0)) %>% 
  group_by(metabolite) %>% 
  mutate(norm_conc = (avg_nmol - min(avg_nmol)) / (max(avg_nmol) - min(avg_nmol))) %>%
  # mutate(norm_conc=(avg_nmol-min(avg_nmol))/(max(avg_nmol)-min(avg_nmol))) %>% 
  # mutate(norm_conc=scale(avg_nmol)[,1]) %>%
  select(metabolite, norm_conc, id) %>%
  pivot_wider(names_from = "metabolite", values_from = "norm_conc") %>%
  mutate(id=factor(id, levels=c(heat_level_order_treat))) %>%
  arrange(id) %>%
  na.omit() %>% 
  column_to_rownames("id")


plotheatmap <- heatmap_data %>%
  data.matrix() %>% 
  heatmaply(fontsize_row = 7, fontsize_col = 7)
  # heatmaply(Rowv = FALSE, fontsize_row = 10, fontsize_col = 7)

plotheatmap
```


### Foldchange K Means clusters?


## Foldchange
```{r}
data_vis_km <- data_vis_qual %>% 
  pivot_wider(names_from = metabolite, values_from = nM) %>% 
  select(-c("treatment", "date")) %>% 
  column_to_rownames("replicate") %>% 
  data.matrix() %>%
  `[<-`(is.na(.), 0) %>%
  scale() %>%
  t() %>%
  kmeans(centers = 6) %>% 
  pluck("cluster") %>%
  as.data.frame() %>%
  set_names("cluster") %>%
  rownames_to_column("metabolite")

data_vis_cluster <- data_vis_fc %>% 
  left_join(data_vis_km) %>% 
  rename(control = C, treatment = name, nM = value) %>% 
  mutate(cluster = factor(cluster))
```


```{r}
data_vis_cluster %>% 
  na.omit() %>% 
  group_by(cluster, date, treatment) %>% 
  filter(str_detect(treatment, "RH|RL|LH|LL")) %>% 
  mutate(avg_fc = mean(fold_change)) %>% 
  ggplot(aes(x = date, y = log2(avg_fc), color = factor(treatment, levels = level_order))) + 
  geom_point() +
  geom_line() + 
  theme_bw() + 
  facet_wrap(~cluster) +
  scale_color_manual(values = c("gold", "lightcoral","orange", "brown3"),
                     labels = c("6N:1P, low dose added", "6N:1P, high dose added",
                                "16N:1P, low dose added", "16N:1P, high dose added"),
                     name = "Treatment") + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5)) + 
  ylab("Average Fold Change") + 
  xlab("Date")

data_vis_cluster %>% 
  na.omit() %>% 
  group_by(cluster, date, treatment) %>% 
  # filter(str_detect(treatment, "RH|RL|LH|LL")) %>% 
  mutate(avg_fc = mean(fold_change)) %>% 
  ggplot(aes(x = date, y = log2(avg_fc), color = factor(treatment, levels = level_order))) + 
  geom_point() +
  geom_line() + 
  theme_bw() + 
  facet_wrap(~cluster) +
  scale_color_manual(values = c("mediumpurple3","darkslategray2", "lightseagreen","gold", "lightcoral","orange", "brown3"),
                     labels = c("0N:1P:1Fe, high dose P","0N:1P, low dose added", "0N:1P, high dose added",
                                "6N:1P, low dose added", "6N:1P, high dose added",
                                "16N:1P, low dose added", "16N:1P, high dose added"),
                     name = "Treatment") + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5)) + 
  ylab("Average Fold Change") + 
  xlab("Date")
```

## Formatting Tote - NEED HELP WITH THIS
I need to make it so that each treatment has a date of 6-27, which will all be the same Tote data
```{r}
# data_vis_form <- data_vis_qual %>% 
#   group_by(metabolite, )
#   pivot_wider(names_from = metabolite, values_from = nM)
```


# General summed metabolite conc
absolute/relative abundances
```{r}
nb.cols <- 19
mycolors <- colorRampPalette(brewer.pal(8, "Paired"))(nb.cols)

metab_type_plot <- data_vis_qual %>% 
  mutate(rate = str_replace(rate, "F", "H")) %>% 
  filter(str_detect(replicate, "Tote|30June|27July")) %>% 
  ggplot() +
  geom_col(aes(x = factor(date), y = nmol_per_pc, fill = metab_type), position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~treatment) + 
  theme_bw()
```

NEED: To put Tote data in each treatment, just as a date. 

Metabolome composition (relative) - 
T0 - lots amino acids, then betaines, then amino acid derivatives, then nucleic acid = sulfur, then purine derivatives.

3 day period - 
C, LH, LL, RH, RL, ZF, ZH, ZL. all have a lot osmolytes - stress??? The fact that all treatments are the same indicates it's not a response to treatment but rather things dying bc of mesocosm? bottle effects?

2 week period - 

4 week period - 
LL, LH, ZF, ZH all have more amino acid derivatives
Osmolytes decrease in all treatments again - indicates adjustment? 
Sulfur metabs - increase in all treatments except ZH? change is less drastic in all treatments apart from LH and RL.

What changes are significant in 

```{r}
nb.cols <- 19
mycolors <- colorRampPalette(brewer.pal(8, "Paired"))(nb.cols)

metab_type_plot <- data_vis_qual %>% 
  mutate(rate = str_replace(rate, "F", "H")) %>% 
  filter(str_detect(replicate, "Tote|30June|14July|27July")) %>% 
  ggplot() +
  geom_col(aes(x = factor(date), y = nmol_per_pc, fill = metab_type), position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~rate) + 
  theme_bw()
```

Trends by rate (relative) - 
nucleoside derivatives: TO - not present. 6/30 - Lots in C, less in L, H. 7/14 and 7/27 - decrease.
Osmolytes - same trend - more gradual die off in C - more stressed? most extreme change in L. (reach equilibrium? less extreme growth than in H?) 
Sulfur - increasing across time with all treatments, fastest increase at 2 week mark but most extreme in C. Least in H (actually decreases at 4 with H)
Amino acid derivatives - 4 week mark greatest in L, but also lots in H. less in C and T0
Betaines - no significant change?

Maybe do something with fold change relative to T0 to see which treatments return closest to "normal environmental conditions"

Trends by rate (absolute) - 
highest absolute nM (not normalized) of summed metabolites in H on week 2. L follows same trend but less extreme. 
highest nM per pc is 6/30? what does this mean.

Overall: 
amino acid relative abundance doesn't change but absolute does. Similar with betaines.
What does this mean?


# QUESTIONS/STORIES - Need those :/

```{r}
sulfur <- data_vis_qual %>% 
  filter(metab_type == "Sulfur") 
  
sulfur %>% 
  group_by(metabolite, treatment, date) %>% 
  # filter(str_detect(metabolite, "Glycerophosphocholine|Phosphocholine")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  ggplot(aes(x = date, y = nM, color = factor(treatment, levels = level_order), group = interaction(date, treatment))) +
  geom_boxplot() + 
  scale_color_manual(values = c("plum2","mediumpurple3","darkslategray2", "lightseagreen","gold", "lightcoral","orange", "brown3"),
                     labels = c("Control","0N:1P:0Fe, high dose P","0N:1P, low dose added", "0N:1P, high dose added",
                                "6N:1P, low dose added", "6N:1P, high dose added",
                                "16N:1P, low dose added", "16N:1P, high dose added"),
                     name = "Treatment") +
  guides(fill=guide_legend(ncol=2)) + 
  # scale_y_log10() + 
  facet_wrap(~metabolite, ncol = 3, scales = "free_y") +
  theme_bw() + 
  xlab("Date") + 
  ylab("nM (nmol per L)")
```

## Osmolytes
```{r}
osmos <- data_vis_qual %>% 
  filter(metab_type == "Osmolyte") 
  
osmos %>% 
  group_by(metabolite, treatment, date) %>% 
  filter(str_detect(metabolite, "Ectoine")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  ggplot(aes(x = date, y = nmol_per_pc, color = factor(treatment, levels = level_order), group = interaction(date, treatment))) +
  geom_boxplot() + 
  scale_color_manual(values = c("plum2","mediumpurple3","darkslategray2", "lightseagreen","gold", "lightcoral","orange", "brown3"),
                     labels = c("Control","0N:1P:0Fe, high dose P","0N:1P, low dose added", "0N:1P, high dose added",
                                "6N:1P, low dose added", "6N:1P, high dose added",
                                "16N:1P, low dose added", "16N:1P, high dose added"),
                     name = "Treatment") +
  guides(fill=guide_legend(ncol=2)) + 
  # scale_y_log10() + 
  facet_wrap(~metabolite, ncol = 2, scales = "free_y") +
  theme_bw() + 
  xlab("Date") + 
  ylab("nM (nmol per L)")
```


# Principal Component Regression