---
title: "peridice_comp"
author: "Iris Kern"
date: "2023-12-14"
output: html_document
---

# Setup
## Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(vegan)
library(heatmaply)
library(ggnewscale)
library(readr)
library(readxl)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(stats)
library(ggfortify)
library(ggh4x)

set.seed(20)
```

data setup
general exploration (heatmaps, nmds, split by metabolite function, pca, metabolite conc abs and rel - is this norm to pc?) 
narrow in on what comes out of that, stories
fun things (machine learning! maybe rfr? predicting pc from metab conc, dissolved related to par?)

Experiment intro

# Data Setup
```{r}
quant_data <- read_csv("csvs/quant_data.csv") %>% 
  filter(nM > 0) %>% 
  filter(nM < 100) 
  
metadataframe <- read_csv("csvs/metadataframe.csv")
pcpn <- read_xlsx("metadata/peridice_pcpn.xlsx") %>% 
  select(tank = Tank, treatment = Treatment2, date = Date, pn = `PN (uM)`, pc = `PC (uM)`, cn = Cnratio,added_N_uM = AddN) %>% 
  mutate(triplicate = str_extract(tank, "\\d")) %>% 
  mutate(tank = str_remove (tank, "1/2|1/3|2/2"))  %>% 
  group_by(tank, date) %>% 
  mutate(pc = mean(pc)) %>% 
  distinct(pc, .keep_all = TRUE) %>% 
  mutate(timepoint = )
  mutate(replicate = paste0("230616_Smp_", ""))
  

data_vis <- quant_data %>%  
  mutate(treatment = str_extract(replicate, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote1|Tote2|Tote3|Tote4")) %>% 
  mutate(timepoint = str_extract(replicate, "27June|30June|14July|21July|27July")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  select(metabolite, treatment, date, nM, replicate) %>% 
  filter(metabolite != "Sodium 2-mercaptoethanesulfonate") %>% 
  filter(metabolite != "Fumaric acid") %>% 
  na.omit() %>% 
  mutate(triplicate = str_extract(replicate, "\\d$")) 

level_order <- c("T0", "C", "ZF", "ZL", "ZH", "LL", "LH", "RL", "RH")

heat_level_order_treat <- c("2022-06-27 Tote", "2022-06-30 C",  "2022-07-14 C",  "2022-07-21 C",  "2022-07-27 C",  
                            "2022-06-30 ZF", "2022-07-14 ZF", "2022-07-21 ZF", "2022-07-27 ZF",
                            "2022-06-30 ZL", "2022-07-14 ZL", "2022-07-21 ZL", "2022-07-27 ZL", 
                            "2022-06-30 ZH", "2022-07-14 ZH", "2022-07-21 ZH", "2022-07-27 ZH",
                            "2022-06-30 LL", "2022-07-14  LL", "2022-07-21 LL", "2022-07-27 LL",
                            "2022-06-30 LH", "2022-07-14 LH", "2022-07-21 LH", "2022-07-27 LH", 
                            "2022-06-30 RL", "2022-07-14 RL", "2022-07-21 RL", "2022-07-27 RL",
                            "2022-06-30 RH", "2022-07-14 RH", "2022-07-21 RH", "2022-07-27 RH")

heat_level_order_date <- c("2022-06-27 Tote", "2022-06-30 C","2022-06-30 ZF", "2022-06-30 ZL", 
                           "2022-06-30 ZH", "2022-06-30 LL","2022-06-30 LH", "2022-06-30 RL", "2022-06-30 RH",
                           "2022-07-14 C", "2022-07-14 ZF", "2022-07-14 ZL", "2022-07-14 ZH", "2022-07-14 LL", 
                           "2022-07-14 LH", "2022-07-14 RL", "2022-07-14 RH", 
                           "2022-07-21 C", "2022-07-21 ZF", "2022-07-21 ZL", "2022-07-21 ZH", 
                           "2022-07-21 LL", "2022-07-21 LH", "2022-07-21 RL", "2022-07-21 RH", 
                           "2022-07-27 C", "2022-07-27 ZF", "2022-07-27 ZL",  
                           "2022-07-27 ZH", "2022-07-27 LL", "2022-07-27 LH", "2022-07-27 RL",  "2022-07-27 RH")

data_vis_norm <- data_vis %>% 
  group_by(replicate, date, metabolite) %>% 
  mutate(nM = mean(nM)) %>% 
  left_join(pcpn) %>% 
  mutate(nmol_per_pc = nM/pc) %>% 
  select(replicate, metabolite, treatment, date, nM, triplicate, nmol_per_pc) %>% 
  mutate(treatment = str_remove(treatment, "\\d"))

metab_groups <- read.csv(
  "https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv",
                              stringsAsFactors = FALSE, header = TRUE) %>% 
  select(metabolite = Compound_Name, emp_form = Empirical_Formula, metab_type = Compound_Type)

```

## Quality Control
```{r}
metab_qual <- read_csv("csvs/peak_quality.csv") %>% 
  select(metabolite = Metabolite, quality = Quality)

data_vis_qual <- data_vis_norm %>% 
  left_join(metab_qual) %>% 
  filter(quality == "Good") %>% 
  mutate(ratio = str_extract(treatment, "R|L|Z|C|Tote")) %>% 
  mutate(rate = str_extract(treatment,  "H$|L$|C|F|Tote")) %>% 
  left_join(metab_groups) %>% 
  filter(metab_type != "Cursed") %>% 
  na.omit()
```

## Metadata
```{r}
pcpn %>% 
  mutate(cn_rank=(cn-min(cn))/(max(cn)-min(cn))) %>%
  filter(str_detect(treatment, "RH|RL|LH|LL|C")) %>% 
  group_by(triplicate, treatment, date) %>%  
  # mutate(avg_cn = mean(cn)) %>% 
  # filter(triplicate == 2) %>% 
  ggplot(aes(x = date, y = cn, group = interaction(treatment, triplicate), color = treatment)) +
  geom_line() + 
  geom_point() + 
  # scale_color_manual(values = c("plum2", "gold", "lightcoral", "darkorange", "brown3"), name = "Treatment") +
  facet_wrap(~treatment) + 
  theme_bw() +  
  ylab("C to N Ratio") + 
  xlab("Date")

pcpn %>% 
  mutate(pc_rank=(pc-min(pc))/(max(pc)-min(pc))) %>%
  filter(!str_detect(treatment, "Tote")) %>% 
  group_by(triplicate, treatment, date) %>%  
  # mutate(avg_cn = mean(cn)) %>% 
  # filter(triplicate == 2) %>% 
  ggplot(aes(x = date, y = pc, group = interaction(treatment, triplicate), color = treatment)) +
  geom_line() + 
  geom_point() + 
  # scale_color_manual(values = c("plum2", "gold", "lightcoral", "darkorange", "brown3"), name = "Treatment") +
  facet_grid(~treatment) + 
  theme_bw() +  
  ylab("Particulate Carbon") + 
  xlab("Date")
```

 

## Stats
### NMDS
```{r}
quant_nmds <- data_vis_qual %>% 
  # arrange(desc(nM)) %>% 
  # group_by(metabolite, replicate) %>% 
  # slice(1) %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nmol_per_pc)) %>%
  ungroup() %>%
  # complete(metabolite, date, treatment) %>%
  # mutate(avg_nmol = replace_na(avg_nmol, 0)) %>%
  mutate(id = paste(date, treatment)) %>% 
  filter(!str_detect(treatment, "Tote"))

quant_mat <- quant_nmds %>%
  group_by(metabolite) %>%
  mutate(norm_conc = rank(avg_nmol)) %>%
  select(metabolite, norm_conc, id) %>% 
  pivot_wider(names_from = "metabolite", values_from = "norm_conc", values_fill = 0) %>%
  column_to_rownames("id") %>%
  data.matrix() 

mdsout <- quant_mat %>%
  metaMDS(k = 2, autotransform = FALSE)

mds_data <- metadataframe %>% 
  ungroup() %>% 
  mutate(timepoint = str_replace_all(timepoint, "July", "-7-22")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "June", "-6-22")) %>% 
  mutate(date = as.Date(timepoint, format = "%d-%m-%y")) %>% 
  filter(str_detect(filename, "Smp")) %>% 
  mutate(treatment = str_remove(treatment, "\\d")) %>% 
  mutate(id = paste(date, treatment))


mdsout$points %>%
  as.data.frame() %>%
  rownames_to_column("id") %>%
  left_join(mds_data) %>%
  mutate(date_fct = factor(date)) %>% 
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=factor(treatment, levels = level_order), shape = date_fct), size=4) +
  scale_color_manual(values = c("plum2","mediumpurple3","darkslategray2", "lightseagreen","gold", "lightcoral","orange", "brown3"),
                     labels = c("Control","0N:1P:1Fe, high dose P","0N:1P, low dose added", "0N:1P, high dose added",
                                "6N:1P, low dose added", "6N:1P, high dose added",
                                "16N:1P, low dose added", "16N:1P, high dose added"),
                     name = "Treatment") +
  scale_shape_discrete(name = "Date") + 
  theme_bw() + 
  ggtitle("Variation of Metabolite Concentration (nM) Across Treatments")
```
Key takeaways: temporal shift but when normalizing to pc, the trends in treatments are more distinct. The controls as well as the 0N treatments remain closer to the 6-30 treatments, but the 6N,16N treatments branch out more.

Cool thoughts: maybe make nmds for ratio/rate? that's normalized to pc?

### PCA
generate pca and check variations explained by each principal component
```{r}
rate_ratio <- data_vis_qual %>% 
  mutate(ratio = str_extract(treatment, "R|L|Z|C|Tote")) %>% 
  mutate(rate = str_extract(treatment,  "H$|L$|C|F|Tote")) %>% 
  filter(!str_detect(treatment, "Tote")) %>% 
  # mutate(rate = str_replace(rate, "F", "H")) %>% 
  mutate(replicate = str_remove(replicate, "230616_Smp_"))

pca_mat <- rate_ratio %>% 
  select(metabolite, replicate, nmol_per_pc) %>% 
  group_by(metabolite) %>% 
  mutate(nmol_per_pc = rank(nmol_per_pc)) %>% 
  pivot_wider(names_from = "metabolite", values_from = "nmol_per_pc") %>% 
  column_to_rownames("replicate") %>% 
  data.matrix() %>% 
  `[<-`(is.na(.), 0)

prcomp(pca_mat) %>%
  pluck("x") %>%
  as.data.frame() %>%
  select(PC1:PC5) %>%
  rownames_to_column("replicate") %>%
  pivot_longer(starts_with("PC"), names_to = "PC", values_to = "PC_val") %>%
  left_join(rate_ratio) %>%
  ggplot() +
  geom_col(aes(x=replicate, y=PC_val, fill=ratio)) +
  # scale_fill_manual(values = c("plum2","gold", "brown3", "lightseagreen"),
  #                    labels = c("Control","6N:1P", "16N:1P", "0N:1P"),
  #                    name = "Treatment") + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5)) + 
  facet_wrap(~PC, ncol=1)

peri_pca_analysis <- prcomp(pca_mat) %>%
  pluck("x") %>%
  as.data.frame() %>%
  select(PC1:PC2) %>%
  rownames_to_column("replicate") %>%
  # pivot_longer(starts_with("PC"), names_to = "PC", values_to = "PC_val") %>%
  left_join(rate_ratio, by = "replicate")  %>% 
  mutate(PC1 = as.numeric(PC1)) %>% 
  mutate(PC2 = as.numeric(PC2))

```

plot pca
```{r}
peri_pca_analysis %>% 
  mutate(ratio = str_replace(ratio, "R", "16N:1P")) %>% 
  mutate(ratio = str_replace(ratio, "L", "6N:1P")) %>%
  mutate(ratio = str_replace(ratio, "Z", "0N:1P")) %>%
  # filter(str_detect(replicate, "21July")) %>% 
  filter(str_detect(rate, "H|L")) %>% 
  mutate(rate = str_replace(rate, "H", "High Dose of Nutrients")) %>% 
  mutate(rate = str_replace(rate, "L", "Low Dose of Nutrients")) %>% 
  ggplot(aes(x = PC1, y = PC2, group = factor(ratio, levels = c("0N:1P", "6N:1P", "16N:1P")))) + 
  # geom_point(aes(fill = ratio, color = ratio), size = 0.5) + 
  geom_point(aes(color = ratio), size = 0.75) + 
  geom_polygon(aes(fill = ratio), alpha = 0.5) + 
  #scale_color_discrete(name = "Ratio") + 
  theme_bw() + 
  facet_grid2(~rate~date, scales = "free", independent = "y")  + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, size = 10), axis.text.y = element_text(size = 10))

peri_pca_analysis %>% 
  mutate(rate = str_replace(rate, "H", "High Dose of Nutrients")) %>% 
  mutate(rate = str_replace(rate, "L", "Low Dose of Nutrients")) %>% 
  mutate(rate = str_replace(rate, "F", "High Dose of Nutrients-no Fe")) %>% 
  # filter(str_detect(replicate, "21July")) %>% 
  filter(str_detect(ratio, "R|L|Z")) %>% 
  # filter(str_detect(rate, "H|L")) %>% 
  mutate(ratio = str_replace(ratio, "R", "16N:1P")) %>% 
  mutate(ratio = str_replace(ratio, "L", "6N:1P")) %>%
  mutate(ratio = str_replace(ratio, "Z", "0N:1P")) %>%
  ggplot(aes(x = PC1, y = PC2, group = rate)) + 
  geom_point(aes(color = rate), size = 0.75) + 
  geom_polygon(aes(fill = rate), alpha = 0.5) + 
  # scale_fill_continuous(values = mycolors) + 
  theme_bw() + 
  facet_grid2(~factor(ratio, levels = c("0N:1P", "6N:1P", "16N:1P"))~date, scales = "free", independent = "y")  + 
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5, size = 7), axis.text.y = element_text(size = 7))
```
Key takeaways: high dose nutrients split up by ratio - become more selective? Low dose nutrients group 16N and 6N together (week 2, nowhere else). Everything shifts at the 4 week mark, metabolome has shifted but what is that shift? More AAs? 

nutrients with 0N never split by rate of nutrient addition. 6N and 16N split by nutrient rate but only really on week 3. This is also again showing a divergence at 4 week mark - shift of metabolome, what is it?

### Heatmap
Grouping - metabolites present at beginning (T0) versus middle (7/14-7/21) versus end (7/27)
```{r}
quant_nmds_heat <- data_vis_qual %>% 
  group_by(metabolite, treatment, date) %>%
  summarise(avg_nmol = mean(nmol_per_pc)) %>%
  ungroup() %>%
  mutate(id = paste(date, treatment)) 

heatmap_data <- quant_nmds_heat %>% 
  complete(metabolite, id, fill = list(avg_nmol = 0)) %>% 
  group_by(metabolite) %>% 
  mutate(norm_conc = rank(avg_nmol)) %>%
  # mutate(norm_conc=(avg_nmol-min(avg_nmol))/(max(avg_nmol)-min(avg_nmol))) %>% 
  # mutate(norm_conc=scale(avg_nmol)[,1]) %>%
  select(metabolite, norm_conc, id) %>%
  pivot_wider(names_from = "metabolite", values_from = "norm_conc") %>%
  mutate(id=factor(id, levels=c(heat_level_order_treat))) %>%
  arrange(id) %>%
  na.omit() %>% 
  column_to_rownames("id")


plotheatmap <- heatmap_data %>%
  data.matrix() %>% 
  heatmaply(fontsize_row = 7, fontsize_col = 7)
  # heatmaply(Rowv = FALSE, fontsize_row = 10, fontsize_col = 7)

plotheatmap
```


### Foldchange K Means clusters?


## General summed metabolite conc
absolute/relative abundances
```{r}
nb.cols <- 19
mycolors <- colorRampPalette(brewer.pal(8, "Paired"))(nb.cols)

data_vis_qual %>% 
  mutate(rate = str_replace(rate, "F", "H")) %>% 
  filter(str_detect(replicate, "Tote|30June|27July")) %>% 
  ggplot() +
  geom_col(aes(x = date, y = nmol_per_pc, fill = metab_type), position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  facet_wrap(~treatment) + 
  theme_bw()
```

NEED: To put Tote data in each treatment, just as a date. 

# QUESTIONS/STORIES - Need those :/

# Principal Component Regression