---
title: "peridice_sandbox"
author: "Iris Kern"
date: `date`
output: html_document
---

GENERAL IDEAS: Want to explore which treatment had most influence on metabolite composition - could do both unknown and known compounds. Could be interesting to explore how time versus treatment alters metabolite composition?

NEED TO WORK ON:
Need to do k means clustering with peak area, can do it with targeted/untargeted combined to see what compounds are behaving similarly to known metabolites.

Quantification step of targeted pipeline

Make stacked barplot with percent peak area in each treatment - group by compound, get total peak area, then split by treatment/date and get percentages of metabolite peak area from that. 

NOTES: get from emily: total POC, fcm, species composition, ATP measurements - emailed

# libraries
```{r}
library(xcms)
library(tidyverse)
library(RaMS)
library(vegan)
library(RColorBrewer)
library(readxl)
library(plotly)
```


# metadataframe, env_data, samp_data, filt_data
```{r}
source('scripts/dataframe_setup.R')
```



# msnexp and peak_data - DON'T RUN UNLESS NECESSARY
```{r}
source('scripts/xcms_peak_output.R')
```

# Normalization by volume filtered
```{r}
peak_data_area_norm <- peak_data_clean %>% 
  left_join(filt_data, by = join_by(timepoint, treatment, replicate)) %>% 
  mutate(area_per_L = area/vol_L_filt) %>% 
  select(-c(`filename.y`)) %>% 
  rename(filename = `filename.x`) %>% 
  mutate(treatment = str_extract(treatment, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "14July", "7-14")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "21July", "7-21")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "30June", "6-30")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "27July", "7-27")) %>% 
  mutate(timepoint = str_replace_all(timepoint, "27June", "6-27")) %>% 
  mutate(date =as.Date(timepoint, format="%m-%d"))
  
chosen_feats <- read_csv("csvs/chosen_feats.csv")

peak_data_norm_good <- peak_data_area_norm %>% 
  filter(samp_type=="Smp") %>% 
  select(compound, filename, area_per_L, mz, rt, treatment, timepoint, date) %>% 
  filter(compound%in%chosen_feats$feature) %>%
  group_by(compound, filename) %>%
  mutate(n=n()) %>%
  group_by(compound) %>%
  filter(all(n==1)) %>%
  ungroup() %>%
  complete(compound, filename, fill=list(area=0)) %>% 
  select(-n)

write.csv(peak_data_norm_good, file = "csvs/untargeted_peak_list.csv", row.names = FALSE)
```

# Normalization to chlorophyll and POC
```{r}

```



# NMDS Plot
## averaged by replicate
```{r}
set.seed(20)

peak_data_nmds <- peak_data_area_norm %>% 
  arrange(desc(area_per_L)) %>% 
  group_by(compound, filename) %>% 
  mutate(filename = str_remove(filename, "\\d.mzML")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  # summarise(n=n()) %>% 
  # filter(n > 1)
  slice(1)

peak_data_nmds <- peak_data_nmds %>% 
  group_by(compound, treatment, timepoint) %>% 
  mutate(avg_area_per_L = mean(area_per_L))
  # distinct(avg_area_per_L)

# peak_data_nmds <- peak_data_nmds %>% 
#   ungroup() %>% 
#   complete(compound, filename) %>% 
#   # filter(is.na(area_per_L)) %>% 
#   mutate(area_per_L = replace_na(area_per_L, 0))

peak_data_nmds <- peak_data_nmds %>% 
  ungroup() %>% 
  group_by(compound, filename) %>% 
  distinct(avg_area_per_L) 

peak_data_nmds <- peak_data_nmds %>% 
  ungroup() %>% 
  # filter(is.na(area_per_L)) %>% 
  complete(compound, filename) %>%
  mutate(avg_area_per_L = replace_na(avg_area_per_L, 0)) %>% 
  mutate(avg_area_per_L = as.character(avg_area_per_L)) 

peri_mat <- peak_data_nmds %>%
  group_by(compound) %>%
  mutate(norm_area=rank(avg_area_per_L)) %>%
  pivot_wider(id_cols = filename, id_expand = TRUE, names_from = "compound", values_from = "norm_area") %>%
  column_to_rownames("filename") %>%
  data.matrix()

# peak_data_nmds %>%
#   filter(compound == "FT1440") %>%
#   ggplot() +
#   geom_col(aes(x = filename, y = area_per_L))
# 
# peri_mat %>%
#   filter(compound == "FT1440") %>%
#   ggplot() +
#   geom_col(aes(x = filename, y = norm_area))

mdsout <- peri_mat %>%
  metaMDS(k = 2, autotransform = FALSE)

mds_data <- metadataframe %>% 
  mutate(filename = str_remove(filename, "\\d.mzML")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  mutate(filename = str_remove(filename, "_"))

mdsout$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(mds_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, label = filename, shape = timepoint), size=4) +
  scale_color_identity(name = "treatment", guide = "legend",labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - No Added Nutrients", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) +
  theme_bw() + 
  ggtitle("NMDS Plot of Entire PERI-DICE Dataset")
```

```{r}
mess_nmds <- peak_data_norm_good %>% 
  arrange(desc(area_per_L)) %>% 
  group_by(compound, filename) %>% 
  mutate(filename = str_remove(filename, "\\d.mzML")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  # summarise(n=n()) %>% 
  # filter(n > 1)
  slice(1) 

mess_nmds <- mess_nmds %>% 
  group_by(compound, treatment, timepoint) %>% 
  mutate(avg_area_per_L = mean(area_per_L))

mess_nmds <- mess_nmds %>% 
  ungroup() %>% 
  # filter(is.na(area_per_L)) %>% 
  complete(compound, filename) %>%
  mutate(avg_area_per_L = replace_na(avg_area_per_L, 0)) %>% 
  mutate(avg_area_per_L = as.character(avg_area_per_L)) 

mess_mat <- mess_nmds %>%
  group_by(compound) %>%
  mutate(norm_area=rank(avg_area_per_L)) %>%
  pivot_wider(id_cols = filename, id_expand = TRUE, names_from = "compound", values_from = "norm_area") %>%
  column_to_rownames("filename") %>%
  data.matrix()

mess <- mess_nmds %>%
  group_by(compound) %>%
  mutate(norm_area=rank(area_per_L)) %>%
  pivot_wider(id_cols = filename, id_expand = TRUE, names_from = "compound", values_from = "norm_area") %>%
  column_to_rownames("filename") %>%
  data.matrix()

messout <- mess %>%
  metaMDS(k = 2, autotransform = FALSE, distance = "euclidean")

mess_data <- metadataframe %>% 
  mutate(filename = str_remove(filename, "\\d.mzML")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  mutate(filename = str_remove(filename, "_"))

messout$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(mess_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, label = filename, shape = timepoint), size=4) +
  scale_color_identity(name = "treatment", guide = "legend", labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - No Added Nutrients", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) +
  scale_shape_discrete(breaks = c("27June", "30June", "14July", "21July", "27July")) +
  theme_bw() + 
  ggtitle("NMDS Plot of Selected Good Peaks")
```

## not averaged by replicate
```{r}
peak_data_nmds <- peak_data_area_norm %>% 
  arrange(desc(area_per_L)) %>% 
  group_by(compound, filename) %>% 
  # summarise(n=n()) %>% 
  # filter(n > 1)
  slice(1) 

peak_data_nmds <- peak_data_nmds %>% 
  ungroup() %>% 
  complete(compound, filename) %>% 
  # filter(is.na(area_per_L)) %>% 
  mutate(area_per_L = replace_na(area_per_L, 0))


peak_data_nmds <- peak_data_nmds %>%
  group_by(compound) %>%
  mutate(norm_area=rank(area_per_L)) %>%
  pivot_wider(id_cols = filename, id_expand = TRUE, names_from = "compound", values_from = "norm_area") %>%
  column_to_rownames("filename") %>%
  data.matrix()

mdsout <- peak_data_nmds %>%
  metaMDS(k = 2, autotransform = FALSE, distance = "euclidean")

mds_data <- metadataframe

mdsout$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(mds_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, label = filename, shape = timepoint), size=4) +
  scale_color_identity(name = "treatment", guide = "legend", labels = c("Redfield Ratio - High N:P", "No N + Low P", 
                                                                        "Low N:P, Low Dose", "T0 - No Added Nutrients", 
                                                                        "Low N:P, High Dose", "No N + High P", "High P, No Fe", 
                                                                        "Redfield Ratio - Low N:P", "Control")) +
  theme_bw() + 
  ggtitle("NMDS Plot of Entire PERI-DICE Dataset")
```


```{r}
mess_nmds <- peak_data_norm_good %>% 
  arrange(desc(area_per_L)) %>% 
  group_by(compound, filename) %>% 
  # summarise(n=n()) %>% 
  # filter(n > 1)
  slice(1) 

mess_nmds <- mess_nmds %>% 
  ungroup() %>% 
  complete(compound, filename) %>% 
  # filter(is.na(area_per_L)) %>% 
  mutate(area_per_L = replace_na(area_per_L, 0))


mess <- mess_nmds %>%
  group_by(compound) %>%
  mutate(norm_area=rank(area_per_L)) %>%
  pivot_wider(id_cols = filename, id_expand = TRUE, names_from = "compound", values_from = "norm_area") %>%
  column_to_rownames("filename") %>%
  data.matrix()

messout <- mess %>%
  metaMDS(k = 2, autotransform = FALSE, distance = "euclidean")

mess_data <- metadataframe

messout$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(mess_data) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, label = filename, shape = timepoint), size=4) +
  scale_color_identity(name = "treatment", guide = "legend", labels = c("Redfield Ratio N:P, High N:P", "No N + Low P",
                                                                        "Low N:P, Low Dose", "T0 - No Added Nutrients",
                                                                        "Low N:P, High Dose", "No N + High P", "No N + High P + No Fe",
                                                                        "Redfield Ratio N:P, Low N:P", "Control")) +
  scale_shape_discrete(breaks = c("27June", "30June", "14July", "21July", "27July")) +
  theme_bw() + 
  ggtitle("NMDS Plot of Selected Good Peaks")

```

# NMDS with targeted data
```{r}
targ_data_qc <- read_csv("csvs/targ_data_qc.csv") %>% 
  mutate(filename = str_remove(filename, "Smp_")) %>% 
  mutate(filename = str_remove(filename, "_\\d"))

filt_data_targ <- filt_data %>% 
  mutate(filename = str_remove(filename, ".mzML")) %>% 
  group_by(treatment, timepoint, replicate)

targ_data <- read_csv("csvs/targeted_peak_list.csv") %>% 
  select(Peptide, Replicate, `Precursor Mz`, `Retention Time`, Area) %>% 
  rename(compound = Peptide) %>% 
  rename(filename = Replicate) %>% 
  rename(mz = `Precursor Mz`) %>% 
  rename(rt = `Retention Time`) %>% 
  rename(area = Area) %>% 
  mutate(replicate = str_extract(filename, "\\d$")) %>% 
  mutate(replicate = str_remove(replicate, ".mzML")) %>% 
  mutate(filename = str_remove(filename, "230616_")) %>% 
  mutate(timepoint = str_extract(filename, "27June|30June|14July|21July|27July")) %>% 
  mutate(treatment = str_extract(filename, "Std|Poo|Blk|C|ZF|ZL|ZH|RL|RH|LL|LH|Tote1|Tote2|Tote3|Tote4")) %>% 
  group_by(treatment, timepoint, replicate)

targ_data <- targ_data %>% 
  left_join(filt_data_targ) %>% 
  mutate(area = as.numeric(area))

targ_data <- targ_data %>% 
  mutate(area_per_L = area/vol_L_filt)
```

```{r}
targ_nmds <- targ_data %>% 
  arrange(desc(area_per_L)) %>% 
  group_by(compound, filename) %>% 
  filter(samp_type == "Smp") %>% 
  mutate(filename = str_remove(filename, "Smp_")) %>% 
  mutate(filename = str_remove(filename, "_\\d$")) %>% 
  # summarise(n=n()) %>% 
  # filter(n > 1)
  slice(1) 

targ_nmds <- targ_nmds %>% 
  ungroup() %>% 
  complete(compound, filename) %>% 
  # filter(is.na(area_per_L)) %>% 
  mutate(area_per_L = replace_na(area_per_L, 0))


targ_mat <- targ_nmds %>%
  group_by(compound) %>%
  mutate(norm_area=rank(area_per_L)) %>%
  pivot_wider(id_cols = filename, id_expand = TRUE, names_from = "compound", values_from = "norm_area") %>%
  column_to_rownames("filename") %>%
  data.matrix()

targout <- targ_mat %>%
  metaMDS(k = 2, autotransform = FALSE, distance = "euclidean")

targ_met <- metadataframe %>% 
  mutate(filename = str_remove(filename, "\\d.mzML")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  mutate(filename = str_remove(filename, "_")) %>% 
  mutate(filename = str_remove(filename, "Blk|Smp|Std"))

targout$points %>%
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(targ_met) %>%
  mutate(col_id=case_when(
    treatment=="C"~"plum2",
    treatment=="ZL"~"darkslategray2",
    treatment=="ZH"~"lightseagreen",
    treatment=="LL"~"gold",
    treatment=="LH"~"lightcoral",
    treatment=="RL"~"orange",
    treatment=="RH"~"brown3",
    treatment=="ZF"~"mediumpurple3", 
    TRUE~"grey50"
  )) %>%
  ggplot() +
  geom_point(aes(x=MDS1, y=MDS2, color=col_id, label = filename, shape = timepoint), size=4) +
  scale_color_identity(name = "treatment", guide = "legend", labels = c("Redfield Ratio N:P, High N:P", "No N + Low P",
                                                                        "Low N:P, Low Dose", "T0 - No Added Nutrients",
                                                                        "Low N:P, High Dose", "No N + High P", "No N + High P + No Fe",
                                                                        "Redfield Ratio N:P, Low N:P", "Control")) +
  scale_shape_discrete(breaks = c("27June", "30June", "14July", "21July", "27July")) +
  theme_bw() + 
  ggtitle("NMDS Plot of Targeted Data")
```


```{r}
biggest_peaks_clean <- peak_data_norm_good %>% 
  group_by(treatment, compound) %>%
  summarise(area_per_L=sum(area_per_L)) %>%
  arrange(desc(area_per_L)) %>% 
  do(head(., n = 20)) %>% 
  na.omit() %>% 
  ungroup()

write.csv(biggest_peaks_clean, file = "biggest_peaks_clean.csv", row.names = FALSE)

biggest_peaks_clean <- read_csv("biggest_peaks_clean.csv")

biggest_peaks_clean_perc <- biggest_peaks_clean %>% 
  group_by(compound) %>% 
  mutate(area_tot = sum(area_per_L)) %>% 
  ungroup() %>% 
  group_by(compound, treatment) %>% 
  mutate(area_perc = area_per_L/area_tot)
 
nb.cols <- 33
mycolors <- colorRampPalette(brewer.pal(8, "Set1"))(nb.cols)

biggest_peaks_clean %>% 
  group_by(treatment, compound) %>%
  summarise(area_per_L=sum(area_per_L)) %>%
  ggplot() +
  geom_col(aes(x=treatment, y=area_per_L, fill = compound), width = 0.5, 
           position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  ggtitle("Metabolite Composition Across Treatment Groups") + 
  theme_bw() + 
  ylab("area per L of sample filtered")

biggest_peaks_clean_perc %>% 
  ggplot() +
  geom_col(aes(x=treatment, y=area_perc, fill = compound), width = 0.5, 
           position = "stack") + 
  scale_fill_manual(values = mycolors) + 
  ggtitle("Metabolite Composition Across Treatment Groups") + 
  theme_bw() + 
  ylab("area per L of sample filtered")
```
percent peak area in each treatment

timeseries of metabolite peak area over time
```{r}
peak_data_norm_good %>% 
  group_by(compound, date, treatment, replicate) %>% 
  mutate(area_avg = mean(area_per_L)) %>% 
  filter(compound == "FT0546") %>% 
  ggplot() +
  # geom_boxplot(aes(x = timepoint, y = area_per_L)) + 
  geom_boxplot(aes(x = as.factor(date), y = area_per_L, fill = treatment)) + 
  scale_fill_manual(values = c("plum2", "lightcoral", "gold", "brown3", "orange", 
                                "grey50", "mediumpurple", "lightseagreen", "darkslategray2")) + 
  theme_bw() +
  scale_y_log10() + 
  xlab("Sample Date") + 
  ylab("Peak Area per Liter Filtered (Log 10 Scale)")

```

# Want to look at Phosphocholine, Glycine Betaine, Homarine
```{r}
# Phosphocholine 

phosphocholine <- peak_data_area_norm %>% 
  mutate(mz, str_extract(mz, "184.073")) %>% 
  na.omit() %>% 
  group_by(date, treatment, ) %>% 
  mutate(area_avg = mean(area_per_L)) %>% 
  filter(rt <= "800") %>% 
  distinct(area_avg)

phosphocholine %>% 
  ggplot() + 
  # geom_boxplot(aes(x = timepoint, y = area_per_L)) + 
  geom_line(aes(x = date, y = area_avg, color = treatment)) + 
  scale_color_manual(values = c("plum2", "lightcoral", "gold", "brown3", "orange", 
                                "grey50", "mediumpurple", "lightseagreen", "darkslategray2")) + 
  theme_bw() +
  xlab("Sample Date") + 
  ylab("Peak Area per Liter Filtered") + 
  ggtitle("Phosphocholine Peak Area Timeseries")
```
phosphocholine is a phospholipid, contains C,N,P,O. Upregulated in LH (High dose N, P). Spike initially in the ZH (no N, high P) but that dies off after 7/14. ZL, C, ZF (all with no P added) never grow very high. RH consistently grows higher. 

# Homarine
```{r}
homarine <- peak_data_area_norm %>% 
  mutate(mz, str_extract(mz, "138.055")) %>%
  na.omit() %>% 
  group_by(treatment, date) %>% 
  mutate(area_avg = mean(area_per_L)) %>%
  filter(rt <= "450") %>% 
  filter(rt >= "300") %>% 
  ungroup() %>% 
  group_by(treatment, date, area_per_L) %>% 
  distinct(area_avg)

ggplot(homarine) + 
  # geom_boxplot(aes(x = timepoint, y = area_per_L)) + 
  geom_line(aes(x = date, y = area_avg, color = treatment)) + 
  scale_color_manual(values = c("plum2", "lightcoral", "gold", "brown3", "orange", 
                                "grey50", "mediumpurple", "lightseagreen", "darkslategray2")) + 
  theme_bw() +
  xlab("Sample Date") + 
  ylab("Peak Area per Liter Filtered") + 
  ggtitle("Homarine Peak Area Timeseries")
```
Homarine - contains N,C,O. Highest in RH (high dose of N,P in Redfield ratio), spikes in C (why), dies in RL, consistent in LL


# Glycine Betaine
```{r}
glycine_betaine <- peak_data_area_norm %>% 
  mutate(mz, str_extract(mz, "118.086")) %>%
  na.omit() %>% 
  select(-c(filename, replicate)) %>% 
  group_by(treatment, date) %>% 
  mutate(area_avg = mean(area_per_L)) %>% 
  filter(rt <= "500") %>%
  filter(rt >= "400") %>% 
  select(c(treatment, date, area_avg)) %>% 
  distinct(area_avg)

ggplot(glycine_betaine) + 
  # geom_boxplot(aes(x = timepoint, y = area_per_L, fill = treatment)) + 
  geom_line(aes(x = date, y = area_avg, color = treatment)) + 
  scale_color_manual(values = c("plum2", "lightcoral", "gold", "brown3", "orange", "grey50", 
                                "mediumpurple", "lightseagreen", "darkslategray2")) + 
  theme_bw() +
  xlab("Sample Date") + 
  ylab("Peak Area per Liter Filtered") + 
  ggtitle("Glycine Betaine Peak Area Timeseries") 
```
Glycine betaine - contains N,C,O. Highest in LH (high N, high P) second highest is RH (Redfield High N,P).

```{r}
glycine_betaine %>% 
  ungroup() %>% 
  # mutate(norm_area = rank(area_per_L)) %>% 
  ggplot() + 
  geom_boxplot(aes(x = timepoint, y = area_per_L, fill = treatment)) + 
  # geom_line(aes(x = date, y = area_avg, color = treatment)) + 
  scale_fill_manual(values = c("plum2", "lightcoral", "gold", "brown3", "orange", "grey50", 
                                "mediumpurple", "lightseagreen", "darkslategray2")) + 
  theme_bw() +
  xlab("Sample Date") + 
  ylab("Peak Area per Liter Filtered") + 
  ggtitle("Glycine Betaine Peak Area Timeseries")
```


